# Sampling

```{r echo=FALSE}
set.seed(123)
```

The generation of data often includes the selection of a subset of
members of a **population**.  This subset is called a **sample**.  
Our focus in statistics is often the **estimation** of population characteristics
(**population parameters**) on the bases of sample characteristics
(**sample statistics**),  

When generating **simulated** datasets we also generate samples.  Simulation
is often done as a means of testing out statistical ideas and methods, 
or (in the case of the bootstrap method) when conventional ways of estimating
standard errors (and therefore confidence intervals) don't work. 

## Random numbers

In both settings (sampling and simulation) we need to start with a method
of generating random numbers.  We'll start by assuming we have 
some means of generating a random sequence of numbers uniformly 
distributed in the interval [0,1]:
\[
    U_i \sim \text{Uniform}(0,1)\qquad \text{for $i=1,2,3, \ldots$}
\]
In R this is done using the `runif()` function.  For example to generate 10 
random draws from Uniform(0,1):
```{r}
runif(10)
```

Another draw will a different set of random numbers:
```{r}
runif(10)
```

If we draw a histogram of 100 draws it looks a bit lumpy:
```{r} 
#| fig.align='center',
#| fig.width=5,
#| fig.cap="100 draws from Uniform(0,1)"
hist(runif(100), xlab="x", ylab="Frequency")
```

But a histogram of 100,000 draws is very uniform:
```{r} 
#| fig.align='center',
#| fig.width=5,
#| fig.cap="100000 draws from Uniform(0,1)"
hist(runif(100000), xlab="x", ylab="Frequency")
```


Although the numbers appear to be random they are in fact generated using
formulae, so if one knows the formula they are predictable.  However the formula
is complex enough that there are no easily discernible patterns in sequences of 
many millions of draws.  These numbers are **psuedo-random**, but they are
random enough for our purposes.  

Underlying the random number generator is a **seed** value, which initiates the
sequence.  If we set the seed to a particular value, the sequence of random
numbers that will be generated after doing so will always be the same.

```{r}
set.seed(111)
runif(10)
runif(10)
runif(10)
```

```{r}
set.seed(8)
runif(10)
```

```{r}
set.seed(111)
runif(10)
```

This property can be very useful when testing algorithms.  