# Statistical Inference

<!--- For HTML Only --- Equivalent code for .pdf must be included in preamble.tex -->
`r if (!knitr:::is_latex_output()) '
$\\DeclareMathOperator*{\\argmin}{argmin}$
$\\newcommand{\\var}{\\mathrm{Var}}$
$\\newcommand{\\bfa}[2]{{\\rm\\bf #1}[#2]}$
$\\newcommand{\\rma}[2]{{\\rm     #1}[#2]}$
$\\newcommand{\\estm}{\\widehat}$'

Let's briefly review a few familiar results from earlier statistics courses.

Most of what we've done in earlier courses has assumed that we have
a **simple random sample** from an **infinite** population.

We'll use data from the `starwars` dataset (in the dplyr package).  There
are data on 87 characters, including their heights in centimetres.  
Six heights are missing, but let's just guess them so we have no missing data
(this process of inventing data has a respectable name:  **'imputation'**, 
and we'll come back to it later).

```{r}
library(dplyr)
data(starwars)
sum(is.na(starwars$height))
starwars <- starwars %>% 
  mutate(height=case_when(name=="Arvel Crynyd" ~ 172, # same as Luke
                          name=="Finn" ~ 172, # same as Luke
                          name=="Rey" ~ 150,  # same as Leia
                          name=="Poe Dameron" ~ 172, # same as Luke
                          name=="BB8" ~ 96, # same as R2D2
                          name=="Captain Phasma" ~ 167, # same as C3P0
                          .default=height))
```

The mean height $\bar{Y}$ of characters in the Star Wars universe is a **population parameter**.
We can **estimate** it from a sample of characters.  For the purposes of this exercise let's imagine 
there are only $N=87$ individuals in the Star Wars universe.

Let's draw a random sample of $n=20$ from this population:
```{r}
set.seed(222)
bigN <- nrow(starwars)
n <- 20
samplemembers <- starwars[sample(bigN, n, replace=FALSE),]
samplemembers %>%
  dplyr::select(name,height,mass,hair_color,homeworld) %>%
  kbl() %>%
  kable_styling()
```

Compute the sample mean $\bar{y}$ and standard deviation $s_y$:
```{r}
ybar <- mean(samplemembers$height)
ybar
sdy <- sd(samplemembers$height)
sdy
```

```{r echo=FALSE}
se <- sdy/sqrt(n)
moe <- 1.96*se
ci <- ybar + moe*c(-1,1)
```

We know how to make a point estimate $\widehat{\bar{Y}}$ for the population
mean $\bar{Y}$ (remember that a hat $\widehat{.}$ over a symbol indicates 
an estimate of that parameter): the population mean is estimated by the sample
mean $\bar{y}$:
\[
    \widehat{\bar{Y}} = \bar{y} = `r sprintf("%.2f",ybar)`
\]
and a confidence interval is given by
\begin{eqnarray}
    \widehat{\bar{Y}} \pm \bfa{MOE}{\widehat{\bar{Y}}}
    &=& \widehat{\bar{Y}} \pm Z\times \bfa{SE}{\widehat{\bar{Y}}}\\
    &=&  \bar{y} \pm Z\times\frac{s_y}{\sqrt{n}}\\
    &=& 
    `r sprintf("%.1f",ybar)` \pm 1.96 \times \frac{`r sprintf("%.1f",sdy)`}{\sqrt{`r n`}}\\
    &=& 
    `r sprintf("%.1f",ybar)` \pm 1.96 \times `r sprintf("%.1f",se)`\\
    &=& 
    `r sprintf("%.1f",ybar)` \pm `r sprintf("%.1f",moe)`\\
    &=& 
    `r sprintf("(%.1f,%.1f)",ci[1],ci[2])`
\end{eqnarray}

