# Sample Designs in R

<!--- For HTML Only --- Equivalent code for .pdf must be included in preamble.tex -->
`r if (!knitr:::is_latex_output()) '
$\\DeclareMathOperator*{\\argmin}{argmin}$
$\\newcommand{\\var}{\\mathrm{Var}}$
$\\newcommand{\\bfa}[2]{{\\rm\\bf #1}[#2]}$
$\\newcommand{\\rma}[2]{{\\rm     #1}[#2]}$
$\\newcommand{\\estm}{\\widehat}$'`

The method of anlalysis of survey data depends crucially on its sample 
design.  Simple methods of analysis which neglect the sample design can
lead to estimates which are severely biased, and standard errors that
are incorrect. 

In the this and the following Chapter we'll use the `survey` package in R to carry out analyses.
We'll use the `schools` data set as our primary example, but also use
data on populations and Gross Domestic Product for each country in 2023
from the 'Our World in Data' website.

```{r echo=FALSE}
#schools <- read.csv("data/nzschools-list.csv")
#names(schools)[c(22,28,39)] <- c("Takiwā","Māori_Electorate","Māori")
#  schools[schools$Regional_Council=="Area Outside Region",]
#schools <- schools[schools$Territorial_Authority!="",]
#schools <- schools[!is.na(schools$Total) & schools$Total>0,]
#schools <- schools |> 
#             mutate(Region=case_when(Territorial_Authority=="Chatham Islands Territory" ~ "Chatham Islands",
#                                 Territorial_Authority=="Rotorua District" ~ "Bay of Plenty Region",
#                                 Territorial_Authority=="Stratford District" ~ "Taranaki Region",
#                                 Territorial_Authority=="Taupo District" ~ "Waikato Region",
#                                 Territorial_Authority=="Waitaki District" ~ "Otago Region",
#                                 .default=Regional_Council))
#write.csv(schools,"data/nzschools.csv",row.names=FALSE)
#is.nested(schools, "Territorial_Authority", "Regional_Council")
#is.nested(schools, "Territorial_Authority", "Region")
```


```{r echo=FALSE}
pop <- read.csv("data/population.csv")
pop <- pop |> filter(Year==2023) |> 
  rename(Country=Entity, Population=all.years) |>
  arrange(Country) 
pop <- pop[!grepl("^OWID_",pop$Code),]
pop <- pop[!grepl("^UN_",pop$Code),]
pop <- pop[pop$Code!="",]
pop$Prob <- pop$Population/sum(pop$Population)
row.names(pop) <- NULL

gdp <- read.csv("data/government-expenditure-vs-gdp.csv")
gdp <- gdp |> filter(Year==2023) |> 
  rename(Country=Entity, Region="World.region.according.to.OWID") |>
  rename(GovtSpend="Government.expenditure") |>
  arrange(Country) 
gdp <- gdp[!grepl("^OWID_",pop$Code),]
gdp <- gdp[!grepl("^UN_",pop$Code),]
gdp <- gdp[gdp$Code!="",]
row.names(gdp) <- NULL

gdp <- merge(gdp,pop)
gdp$Prob <- gdp$Population/sum(gdp$Population)
gdp$GDPPerCapita <- gdp$GDP/gdp$Population

mil <- read.csv("data/military-spending-sipri.csv")
#mil <- read.csv("data/military-expenditure-as-a-share-of-government-spending.csv")
mil <- mil |> filter(Year==2023) |> 
  rename(Country=Entity, Region="World.region.according.to.OWID") |>
  rename(MilSpend=Military.expenditure) |>
  select(-Region) |>
  arrange(Country) 
mil <- mil[!grepl("^OWID_",pop$Code),]
mil <- mil[!grepl("^UN_",pop$Code),]
mil <- mil[pop$Code!="",]
row.names(mil) <- NULL

mil <- merge(gdp,mil)
```

```{r echo=FALSE}
schools <- read.csv("data/nzschools.csv")
```

```{r}
# Census
schools.census <- schools |> select.srswor(n=nrow(schools))
# SRSWOR
schools.srswor <- schools |> select.srswor(n=20)
# PPSWR
schools.ppswr <- schools |> select.ppswr(n=20, sizevar="Total")
# STSRS
schools.stsrs <- schools |> select.stsrs(n=60, stratumvar="Education_Region")
# 1SC
schools.1sc <- schools |> select.1sc(n=10, clustervar="Territorial_Authority")
# 2SC
schools.2sc <- schools |> select.2sc(n=10, cfraction=0.20, mmin=2, 
                                     clustervar="Territorial_Authority")
# ST2SC
schools.st2sc <- schools |> select.st2sc(n=30, stratumvar="Region", allocation="proportional",
                                cfraction=0.20, clustervar="Territorial_Authority")
```

In order to make estimates from sampled data using the `survey` package
we need to specify the sample design we're using. 
We'll do this for the various sample designs covered in the 
previous chapter.  

> Note that in the `survey` package we identify variables with their names without
> quotes and with a `~` symbol in front.  (If you're interested you can 
> use the `survyr` package which is a `dplyr`-like wrapper for `survey` 
> with functions defined so that the `~` isn't necessary.)

We'll define each design, and then use it to estimate the mean 
size of school rolls in the population (stored in the variable `Total` in
the `schools` data frame.  (One exception will be the PPSWR example
where we'll select a sample of countries using their population size
as the size variable.)

The true mean school size is
```{r}
ybar <- mean(schools$Total)
ybar
```

## Census

The analysis of a census is trivial (at least, if there is no non-response).
We can create a census dataset by taking an SRSWOR of size $n=N$.
```{r}
schools.census <- schools |> select.srswor(n=nrow(schools))
```
This function adds the column `bigN` to the dataset, which contains 
population size $N$ repeated on every row.

Now we can specify the design.
```{r}
schools.census.des <- svydesign(id=~1, data=schools.census, fpc=~bigN)
```
Note that we tell R which column contains the population size using the `fpc` argument. 
Our data aren't clustered, so the `id` variable (which identifies clusters) is set to 1.

```{r}
svymean(~Total, design=schools.census.des, deff=TRUE)
```
The estimated mean is `r sprintf("%.2f", ybar)`, which is of course
exactly equal to the population mean - and note that the standard error is zero:
censuses have no sampling error.  We've measured everything there is to measure
so we have no uncertainty.  

We've requested `Deff=TRUE`, and have got a Deff value of NaN, which means
not able to be calculated.  We'll discuss what this means later on.  

If we had left out the `fpc` command 
```{r}
schools.census.des <- svydesign(id=~1, data=schools.census)
```
we get a warning that we can interpret as R assuming an SRSWR from an infinite population.

The estimated mean is the same, but the standard error are now what'd we'd have learned 
from first year statistics.
```{r}
svymean(~Total, design=schools.census.des, deff=TRUE)
```
```{r}
sy <- sd(schools.census$Total)
n <- nrow(schools.census)
se <- sy/sqrt(n)
```

i.e. the standard error is
\[
   \bfa{SE}{\widehat{\bar{Y}}} 
   = \frac{s_y}{\sqrt{n}} 
   = \frac{`r sprintf("%.3f",sy)`}{\sqrt{`r n`}} = `r sprintf("%.3f",se)`
\]

Also notice that the `Deff` value has changed to `Inf` (meaning infinity) -
again we'll come back to what this means this shortly.  

## Simple Random Sample WithOut Replacement (SRSWOR)

```{r echo=FALSE}
set.seed(111)
```

Let's draw a sample of $n=60$ schools from the population by SRSWOR.
```{r}
schools.srswor <- schools |> select.srswor(n=60)
```

To specify the design we use the same syntax as above:
```{r}
schools.srswor.des <- svydesign(id=~1, data=schools.srswor, fpc=~bigN)
```

Estimating the mean school size is also just the same:
```{r}
svymean(~Total, design=schools.srswor.des, deff=TRUE)
```

This is the mean of a genuine sample now, so we have an estimated mean value that
differs from the truth, and the standard error is gauge of its accuracy.

The standard error comes from the finite population corrected formula:
```{r}
bigN <- first(schools.srswor$bigN) # take the entry in the first row: they're all the same
sy <- sd(schools.srswor$Total)
n <- nrow(schools.srswor)
se <- sqrt(1-n/bigN)*sy/sqrt(n)
```
\[
   \bfa{SE}{\widehat{\bar{Y}}} 
   = \sqrt{1-\frac{n}{N}}\frac{s_y}{\sqrt{n}} 
   = \sqrt{1-\frac{`r n`}{`r bigN`}}
     \frac{`r sprintf("%.3f",sy)`}{\sqrt{`r n`}} = `r sprintf("%.3f",se)`
\]

Compare this to the simple SE, which is just slightly bigger
```{r}
se <- sy/sqrt(n)
```
\[
   \bfa{SE}{\widehat{\bar{Y}}} 
   = \frac{s_y}{\sqrt{n}} 
   = \frac{`r sprintf("%.3f",sy)`}{\sqrt{`r n`}} = `r sprintf("%.3f",se)`
\]
This is what we get if we omit the `fpc`
```{r}
schools.srswor.des <- svydesign(id=~1, data=schools.srswor)
svymean(~Total, design=schools.srswor.des, deff=TRUE)
```

## Probability Proportional to Size With Replacement (PPSWR)

We have a dataset from 'Our World in Data' giving population sizes, 
Gross Domestic Product and military expenditure in 2023 (all in US dollars).
The first 10 rows are below:
```{r echo=FALSE}
mil |>
  slice_head(n=10) |>
  select(-Year) |>
  select(-Prob) |>
  kbl() |>
  kable_styling()
```

We can select a PPSWR using population as the size variable
```{r}
mil.ppswr <- mil |> select.ppswr(n=30, sizevar="Population")
```

The `weight` column here combines the weight over the number of times a unit is selected
```{r}
mil.ppswr.des <- svydesign(id=~1, data=mil.ppswr, weight=~weight)
```

The mean military expenditure per country is
```{r}
svymean(~MilSpend, design=mil.ppswr.des, deff=TRUE)
```
though more interesting is the **total** of this variable, to see
how much the world is spending.
```{r}
svytotal(~MilSpend, design=mil.ppswr.des, deff=TRUE)
```

Note again that in a with replacement design we can't apply the finite
population correction.  

## Stratified Sample

```{r echo=FALSE}
set.seed(112)
```

Recall that in this design we specify a set of groups called strata, and
each population unit belongs to one of these strata.

Selecting a sample of $n=60$ schools by equal allocation across the 12 
Education Regions:
```{r}
schools.stsrs <- schools |> select.stsrs(n=60, stratumvar="Education_Region")
```

Specify this design
```{r}
schools.stsrs.des <- svydesign(id=~1, strata=~Education_Region, fpc=~Nh,
                               data=schools.stsrs,
                               survey.lonely.psu="remove")
```
Note that we've included the option `survey.lonely.psu="remove"` to mean
that the estimation of standard errors doesn't fail if a stratum only has a single
PSU.  

We can now return to estimating the mean school size
```{r}
svymean(~Total, design=schools.stsrs.des, deff=TRUE)
```

Now it's finally time to talk about the **Design Effect**, or the Deff for short.

The Deff of an estimator $\widehat{T}$ of a population characteristic $T$ in
a particular sample design is defined to be
\[
   \bfa{Deff}{\widehat{T}_{\text{design}}}
   = 
   \frac{\bfa{Var}{\widehat{T}_{\text{design}}}}{
         \bfa{Var}{\widehat{T}_{\text{SRSWOR}}}}
   = 
   \left(\frac{\bfa{SE}{\widehat{T}_{\text{design}}}}{
               \bfa{SE}{\widehat{T}_{\text{SRSWOR}}}}\right)^2
\]
where both estimators are based on samples of the same size.

It tells us how the standard error differs between our design of 
interest and a SRSWOR of the same sample size.

Note that

* We have a different Deff for every estimate we make from a survey.
* A Deff$<$1 means that the Standard Errors are smaller than what
  we'd get in SRSWOR.  A small Deff indicates a **more efficient** design.
* A Deff of 1 means the design is no better than SRSWOR - and of course
  (look back above) the Deffs of all estimates in SRSWOR are 1 by definition.
* If we're dealing with a census, then the Standard Error is zero, 
  and in any case there's no difference between designs, so the Deff
  isn't defined.
* **Stratified designs** can achieve Deffs much much less than 1 if their strata
  are well defined to be informative about the estimates of interest.  e.g.
  in business surveys where the strata are defined by business size we 
  can get very good Deffs.
* **Cluster designs** often have Deffs greater than 1, which show them to be 
  significantly less efficient than SRSWOR.
  
The Deff values we get from R are approximations, and aren't very reliable
when sample sizes are small.  But do notice that we have a Deff<1 here
for the stratified sample of $n=60$ units, and the SE is lower than
that achieved in the SRSWOR of $n=60$ units above.
  
```{r echo=FALSE}
#(SE(svymean(~Total, design=schools.stsrs.des, deff=TRUE))/
#  SE(svymean(~Total, design=schools.srswor.des, deff=TRUE)))^2
```

## One stage cluster design (1SC)

```{r echo=FALSE}
set.seed(221)
```

A one stage cluster design takes $n=10$ Territorial Authority areas (TAs) at random
and selects all schools in each selected TA.
```{r}
schools.1sc <- schools |> select.1sc(n=10, clustervar="Territorial_Authority")
```
We need a column `bigN` which gives the number of TAs in the population.  

```{r}
schools.1sc.des <- svydesign(id=~Territorial_Authority, data=schools.1sc, 
                             fpc=~bigN)
```

```{r}
svymean(~Total, design=schools.1sc.des, deff=TRUE)
```
The Deff of this one stage cluster sample is bad (much larger than 1) - this is due to 
strong differences between TAs in the mean school size.  The more the clusters differ from
each other on the characteristic of interest, the worse the Deff gets.

## Two stage cluster design (2SC)

```{r echo=FALSE}
set.seed(222)
```

A two stage cluster design takes $n=10$ Territorial Authority areas (TAs) at random
and selects 20% of the schools in each selected TA.
```{r}
schools.2sc <- schools |> select.2sc(n=10, cfraction=0.20, mmin=2, 
                                     clustervar="Territorial_Authority")
```
We need a column `bigN1` which gives the number of TAs in the population, and
`bigN2` which is the number of schools in each TA.  We also need to specify names
of unique identifiers of the first stage units (the PSUs: these are Territorial Authorities)
and the second stage units (the SSUs: these are Schools).

```{r}
schools.2sc.des <- svydesign(id=~Territorial_Authority+School_Id, data=schools.2sc, 
                             fpc=~bigN1+bigN2,
                             survey.lonely.psu="remove")
```

```{r}
svymean(~Total, design=schools.2sc.des, deff=TRUE)
```

## Stratified two stage cluster design

```{r echo=FALSE}
set.seed(223)
```

Finally here is a stratified two stage sample of schools within TA's with Regions.
The Regions are strata, the TAs are clusters within the Regions, and then 20% of 
schools are sampled in the selected TAs.
```{r}
schools.st2sc <- schools |> select.st2sc(n=30, stratumvar="Region", allocation="proportional",
                                         cfraction=0.20, clustervar="Territorial_Authority")
```

The design is the same as the two stage cluster sample, except we add the stratum specification.
```{r}
schools.st2sc.des <- svydesign(id=~Territorial_Authority+School_Id, strata=~Region, data=schools.st2sc, 
                               fpc=~bigN1+bigN2,
                               survey.lonely.psu="remove")
```

```{r}
svymean(~Total, design=schools.st2sc.des, deff=TRUE)
```





