# Survey Analysis

<!--- For HTML Only --- Equivalent code for .pdf must be included in preamble.tex -->
`r if (!knitr:::is_latex_output()) '
$\\DeclareMathOperator*{\\argmin}{argmin}$
$\\newcommand{\\var}{\\mathrm{Var}}$
$\\newcommand{\\bfa}[2]{{\\rm\\bf #1}[#2]}$
$\\newcommand{\\rma}[2]{{\\rm     #1}[#2]}$
$\\newcommand{\\estm}{\\widehat}$'`

With the ability to load a survey dataset and specify its survey design
we're now ready to do some analysis.  

We'll use the `api` dataset which comes with the survey package: it's a
set of observations of the Academic Performance Index (API) on all 
California Schools with at least 100 students in the years 1999 and 2000.  
```{r}
data(api)
```

```{r echo=FALSE}
apivarnames <- read.csv("data/apivarnames.csv")
```

In particular we'll use the `apiclus2` dataset, which is a two stage
cluster sample of 126 schools drawn from the population of 6194.
The top 10 rows of the data set and some of the key variables are shown below.
```{r}
apiclus2$stype <- factor(apiclus2$stype, levels=c("E","M","H"))
levels(apiclus2$stype) <- c("Elementary","Middle","High")

apiclus2 |>
  slice_head(n=10) |>
  select(cds, stype, name, dnum, cnum, api99, api00) |>
  kbl() |>
  kable_styling()
```

The full set of variables in the dataset is:
```{r echo=FALSE}
apivarnames |>
  kbl(col.names=c("Variable","Description")) |>
  kable_styling(full_width=FALSE)
```

The `apiclus2` dataset was drawn by taking a sample of districts (labelled by district number `dnum`),
and then a sample of schools within districts (schools are labelled by their school number `snum`).  
The column `fpc1` contains the number of districts (757), and the column `fpc2` contains the
number of schools in each district (which varies by district).

We can specify this two stage cluster design using
```{r}
apiclus2.des <- svydesign(ids=~dnum+snum, fpc=~fpc1+fpc2, data=apiclus2)
```

The `summary()` function tells us the basics of the design we've just specified:
```{r}
summary(apiclus2.des)
```

We can see that from the summary that $n=40$ districts were sampled (out of $N=757$).
Let's count the number of schools sampled per district.  We'll just display the top 15 largest 
districts:
```{r message=FALSE}
apiclus2 |>
  group_by(dnum,dname) |>
  summarise(Mi=first(fpc2),mi=n()) |>
  ungroup() |>
  arrange(desc(Mi)) |>
  slice_head(n=15) |>
  kbl(col.names=c("dnum","District Name","$M_i$","$m_i$"),escape=FALSE) |>
  kable_styling(full_width=FALSE)
```
After the 10 largest districts a census of schools is taken in each selected
district $m_i=M_i$

## Continous variables

Consider the key outcome variable `api00``, the API score in the year
2000 for each school.  

### Display {-}

The basic workhorse for the display of numerical variables is the histogram
```{r}
#| fig.width=5,
#| fig.align="center",
#| fig.cap="Weighted histogram of API (2000) scores"
svyhist(~api00, design=apiclus2.des, prob=FALSE,
        xlab="API Score (Year 2000)", 
        main="Weighted distribution of API (2000) scores")
```

Note that this histogram differs strongly from the simple histogram
of the sample values:
```{r}
#| fig.width=5,
#| fig.align="center",
#| fig.cap="Unweighted histogram of API (2000) sample data"
hist(apiclus2$api00, 
        xlab="API Score (Year 2000)", 
        main="Unweighted distribution of API (2000) scores")
```
a warning to us that with survey data we can go badly astray if we ignore 
the sample design.

Boxplots are an alternative to histograms:
```{r}
#| fig.width=5,
#| fig.align="center",
#| fig.cap="Distribution of API (2000) sample data"
svyboxplot(api00~1, design=apiclus2.des, 
           ylab="API (2000)", main="API (2000) Scores")
```

... especially if we want to display variables divided by a category: 
```{r}
#| fig.width=5,
#| fig.align="center",
#| fig.cap="Distribution of API (2000) sample data"
svyboxplot(api00~stype, design=apiclus2.des, 
           xlab="School Type",
           ylab="API (2000)", main="API (2000) Scores")
```

If we have two continuous variables we can plot them as a scatterplot, showing
the weights of each observation:
```{r}
#| fig.width=5,
#| fig.align="center",
#| fig.cap="Distribution of API scores in 1999 and 2000"
svyplot(api00~api99, design=apiclus2.des, style="bubble",
        xlab="API (1999)", ylab="API (2000)")
```

### Estimation {-}

We've already seen that we can estimate means using the `svymean()` function.
```{r}
svymean(~api00, design=apiclus2.des)
```
The `jtools` package allows us to estimate the population standard deviation too
```{r}
library(jtools)
svysd(~api00, design=apiclus2.des)

```
(Pause a minute and remind yourself of the difference between a standard deviation and
a standard error.)

We can estimates means for a set of separate variables, using a perhaps uninintuitive notation:
```{r}
svymean(~api99+api00, design=apiclus2.des)
```
or for a variable split by categories by wrapping `svymean()` up inside a call to `svyby()`
```{r}
svyby(~api00, by=~stype, svymean, design=apiclus2.des)
```

A **dotchart** is a nice way of showing the estimated means and their confidence intervals:
```{r}
#| fig.width=5,
#| fig.align="center",
#| fig.cap="Mean API (2000) scores by School Type"
ss <- svyby(~api00, by=~stype, svymean, design=apiclus2.des)
sc <- confint(ss)
ns <- nrow(sc)
dotchart(ss, xlim=range(sc), xlab="Mean API (2000) Score")
arrows(sc[,1], 1:ns, sc[,2], 1:ns, angle=90, code=3, length=0.1)
```

If we want to test statistical hypotheses about differences between groups
one way is to use the regression framework
```{r}
summary(svyglm(api00~stype, design=apiclus2.des))
```
Here we can see that with Elementary as the reference level we 
get the same estimate of the mean API reported as the `(Intercept)`.
We further see that there is no significant difference between Middle
and Elementary Schools in mean API, but the mean API is significantly lower
at High Schools compared to Elementary Schools.

Regressions take of course take lots of variables into account, and
we can apply familiar regression techniques to investigating 
conditional associations between predictors and a continuous outcome. 

For example we can test whether the percentage of children at a school
who have free school meals (i.e. the `meals` variable) associates with the 
API score:
```{r}
summary(svyglm(api00~stype+meals, design=apiclus2.des))
```
... and there is indeed a strong negative association: the higher the
`meals` value the lower the API score.

We can estimate **totals** using the `svytotal()` function - for example
we can estimate the total number of students enrolled at all schools in 
California:
```{r}
svytotal(~enroll, design=apiclus2.des, na.rm=TRUE)
```
Although note that we have had to remove some schools with unknown
`enroll` values by stating `na.rm=TRUE`, and these forced removals likely mean 
that our estimate of the total is an underestimate.

Confidence intervals can be calculated using the `confint()` function
```{r}
confint(svymean(~api99+api00, design=apiclus2.des), level=0.95)
```

```{r}
confint(svyby(~api00, by=~stype, svymean, design=apiclus2.des))
```

The standard errors can be directly extracted using `SE()`
```{r}
SE(svyby(~api00, by=~stype, svymean, design=apiclus2.des))
```
The **relative standard errors**, also known as **relative sampling errors** 
(RSEs) can be extracted by `cv()`:
```{r}
cv(svyby(~api00, by=~stype, svymean, design=apiclus2.des))
```
RSEs are a useful measure of quality of estimates of quantities that are 
strictly positive, such as the API score.

```{r echo=FALSE}
ss <- svymean(~api00, design=apiclus2.des)
cc <- cv(ss)
```

If we learned that the SE of an estimate (e.g. the mean of `api00`) was 
`r round(as.numeric(SE(ss)),2)` we wouldn't be able to assess whether it 
was a good estimate or not without knowing that the value of the estimate
was `r round(as.numeric(ss),2)`.  A good way to assess the quality of 
an estimate is the SE of an estimate $\widehat{T}$ as a proportion of 
the estimate $\widehat{T}$:
\[
   \bfa{RSE}{\widehat{T}} = \frac{\bfa{SE}{\widehat{T}}}{\widehat{T}}
\]
For overall mean `api00` the RSE is
\[
   \bfa{RSE}{\widehat{\bar{Y}}} 
   = \frac{`r round(as.numeric(SE(ss)),2)`}{`r round(as.numeric(ss),2)`} 
   = `r round(as.numeric(cc),3)`
\]
This is a highly precise estimate: the SE is only `r round(100*as.numeric(cc),1)`%
of the value of the estimate.

Good RSEs are only a few percent at most. Poor RSEs are larger than 10%, and anything bigger
than 20% isn't really publishable.

In general a standard deviation divided by a mean goes by the term
**coefficient of variation** (CV), which is why the R function is 
called `cv()`.
```{r}
cv(svymean(~api00, design=apiclus2.des))
```

## Categorical Variables

The workhorse display for categorical variables is the barplot, 
showing **proportions** in categories.  Proportions are first calculated
using the `svymean()` function:
```{r}
#| fig.cap="Estimates of the proportions of Schools of each type",
#| fig.align="center",
#| fig.width=5
barplot(svymean(~stype, design=apiclus2.des),
        names=levels(apiclus2$stype),
        xlab="School Type", ylab="Proportion")
```

```{r}
svymean(~stype, design=apiclus2.des)
```

It's worth pausing to recognise that a proportion is a mean of a special
type of variable called an **indicator variable**: it takes the value 1 if
the observation is a member of specific category and 0 if not.  Add up all the ones 
and zeroes and you get the number of observations in the category, and divide
by the number of observations you get a proportion.
\[
    p = \frac{1 + 0 + 1 + 1 + 0 + 0 + ....}{n} = \frac{1}{n}\sum_{i=1}^n Y_i
\]
where $Y_i$ is an indicator variable.

With sample data we calculate **weighted sums** though:
\[
   \widehat{p} = \frac{\sum_{i=1}^n w_i Y_i}{\sum_{i=1}^n w_i} = \frac{\widehat{Y}}{\widehat{N}}
\]
where the top line of this fraction is $\widehat{Y}$, our estimate of the
number of individuals (in a finite population) who belong to the category of
interest.  The bottom line is $\widehat{N}$, the estimated size of the population.
$\widehat{N}$ is exact in some designs (e.g. SRSWOR, or Stratified SRSWOR), but
is approximate in other designs (especially cluster designs).  

These total estimates within categories can be estimated using the `svytotal()`
function.  
```{r}
svytotal(~stype, design=apiclus2.des)
```

When creating barplots of estimated proportions and/or totals 
we really should indicate the uncertainties in our
estimates, and show confidence intervals on the bars.  We can use the standard
errors provided by `svymean()` and `svytotal()`: the function `confint()` 
will provide these:
```{r}
confint(svytotal(~stype, design=apiclus2.des))
```
These confidence intervals are symmetric about the point estimates:
```{r}
#| fig.cap="Estimated proportions of schools by school type - with symmetric confidence intervals",
#| fig.width=5,
#| fig.align="center"
ss <- svymean(~stype, design=apiclus2.des)
ci <- confint(ss)
bp <- barplot(ss, xlab="School Type", ylab="Estimated Size",
              names=levels(apiclus2$stype),
              ylim=c(0,max(ci)))
arrows(bp, ci[,1], bp, ci[,2], angle=90, length=0.2, code=3)
```

```{r tabci.symmetric}
names(ss) <- levels(apiclus2$stype)
cbind(as.numeric(ss),ci) |>
  kbl(col.names=c("Estimate","2.5%","97.5%"), 
      digits=3,
      caption="Symmetric confidence interval estimates") |>
  kable_styling(full_width=FALSE)
```

These symmetric intervals are actually somewhat unrealistic, because 
`confint()` doesn't know it's dealing with indicator variables, and
doesn't take into account the fact that proportions and totals can't go 
negative.  

A better method of constructing confidence intervals is the special
function `svyciprop()`: 
```{r}
svyciprop(~I(stype=="Elementary"), design=apiclus2.des)
```
Note that we've had to create an indicator variable here using the `I()`
function for just one level in order to calculate its confidence interval.

To get a confidence interval proportion for every level of a factor 
it's somewhat fiddly with the raw `survey` package:
```{r}
#| fig.cap="Estimated proportions of schools by school type - with asymmetric confidence intervals",
#| fig.width=5,
#| fig.align="center"
ss <- svymean(~stype, design=apiclus2.des)
ci <- t(sapply(levels(apiclus2$stype),
       function(stypelevel) {
          form <- eval(parse(text=paste0("formula(~I(stype==\"",stypelevel,"\"))")))
          confint(svyciprop(form, design=apiclus2.des))    
       }))
bp <- barplot(ss, xlab="School Type", ylab="Estimated Size",
              names=levels(apiclus2$stype),
              ylim=c(0,max(ci)))
arrows(bp, ci[,1], bp, ci[,2], angle=90, length=0.2, code=3)
```
```{r tabci.asymmetric}
names(ss) <- levels(apiclus2$stype)
cbind(as.numeric(ss),ci) |>
  kbl(col.names=c("Estimate","2.5%","97.5%"), 
      digits=3,
      caption="Asymmetric confidence interval estimates") |>
  kable_styling(full_width=FALSE)
```

The biggest thing to notice is that the confidence interval for "High"
is now asymmetric: it's longer on the high side that the low side, and
more correctly represents our uncertainty.  The estimates for "Elementary"
and "Middle" are essentially unchanged.

For **crosstabulations** we can use `svytable()`, which gives weighted
counts in categories - e.g. of whether School Wide growth targest had been met
```{r}
svytable(~stype+sch.wide, design=apiclus2.des)
```
A test of association is given by the weighted Chi-Square test
```{r}
svychisq(~stype+sch.wide, design=apiclus2.des)
```
We can carry out a logistic regression to make a similar test of association.  
We have `sch.wide` as the binary outcome variable and `stype` as the predictor:
```{r}
summary(svyglm(I(sch.wide=="Yes")~stype, family=quasibinomial(link="logit"), 
               design=apiclus2.des))
```

With regular `lm()` and `glm()` regression fits we can use F tests to determine
whether specific terms in the model should be included.  A similar test
can be done to test for the inclusion of specific terms using `regTermTest` 
```{r}
sr <- svyglm(I(sch.wide=="Yes")~stype, family=quasibinomial(link="logit"), 
               design=apiclus2.des)
regTermTest(sr, ~stype)
```
Here we can see that `stype` is contributing significantly to the fit of the model. 
