# Survey Analysis

<!--- For HTML Only --- Equivalent code for .pdf must be included in preamble.tex -->
`r if (!knitr:::is_latex_output()) '
$\\DeclareMathOperator*{\\argmin}{argmin}$
$\\newcommand{\\var}{\\mathrm{Var}}$
$\\newcommand{\\bfa}[2]{{\\rm\\bf #1}[#2]}$
$\\newcommand{\\rma}[2]{{\\rm     #1}[#2]}$
$\\newcommand{\\estm}{\\widehat}$'`

With the ability to load a survey dataset and specify its survey design
we're now ready to do some analysis.  

We'll use the `api` dataset which comes with the survey package: it's a
set of observations of the Academic Performance Index (API) on all 
California Schools with at least 100 students in the years 1999 and 2000.  
```{r}
data(api)
```

```{r echo=FALSE}
apivarnames <- read.csv("data/apivarnames.csv")
```

In particular we'll use the `apiclus2` dataset, which is a two stage
cluster sample of 126 schools drawn from the population of 6194.
The top 10 rows of the data set and some of the key variables are shown below.
```{r}
apiclus2$stype <- factor(apiclus2$stype, levels=c("E","M","H"))
levels(apiclus2$stype) <- c("Elementary","Middle","High")

apiclus2 |>
  slice_head(n=10) |>
  select(cds, stype, name, dnum, cnum, api99, api00) |>
  kbl() |>
  kable_styling()
```

The full set of variables in the dataset is:
```{r echo=FALSE}
apivarnames |>
  kbl(col.names=c("Variable","Description")) |>
  kable_styling(full_width=FALSE)
```

The `apiclus2` dataset was drawn by taking a sample of districts (labelled by district number `dnum`),
and then a sample of schools within districts (schools are labelled by their school number `snum`).  
The column `fpc1` contains the number of districts (757), and the column `fpc2` contains the
number of schools in each district (which varies by district).

We can specify this two stage cluster design using
```{r}
apiclus2.des <- svydesign(ids=~dnum+snum, fpc=~fpc1+fpc2, data=apiclus2)
```

The `summary()` function tells us the basics of the design we've just specified:
```{r}
summary(apiclus2.des)
```

We can see that from the summary that $n=40$ districts were sampled (out of $N=757$).
Let's count the number of schools sampled per district.  We'll just display the top 15 largest 
districts:
```{r message=FALSE}
apiclus2 |>
  group_by(dnum,dname) |>
  summarise(Mi=first(fpc2),mi=n()) |>
  ungroup() |>
  arrange(desc(Mi)) |>
  slice_head(n=15) |>
  kbl(col.names=c("dnum","District Name","$M_i$","$m_i$"),escape=FALSE) |>
  kable_styling(full_width=FALSE)
```
After the 10 largest districts a census of schools is taken in each selected
district $m_i=M_i$

## Continous variables

Consider the key outcome variable `api00``, the API score in the year
2000 for each school.  

### Display {-}

The basic workhorse for the display of numerical variables is the histogram
```{r}
#| fig.width=5,
#| fig.align="center",
#| fig.cap="Weighted histogram of API (2000) scores"
svyhist(~api00, design=apiclus2.des, prob=FALSE,
        xlab="API Score (Year 2000)", 
        main="Weighted distribution of API (2000) scores")
```

Note that this histogram differs strongly from the simple histogram
of the sample values:
```{r}
#| fig.width=5,
#| fig.align="center",
#| fig.cap="Unweighted histogram of API (2000) sample data"
hist(apiclus2$api00, 
        xlab="API Score (Year 2000)", 
        main="Unweighted distribution of API (2000) scores")
```
a warning to us that with survey data we can go badly astray if we ignore 
the sample design.

Boxplots are an alternative to histograms:
```{r}
#| fig.width=5,
#| fig.align="center",
#| fig.cap="Distribution of API (2000) sample data"
svyboxplot(api00~1, design=apiclus2.des, 
           ylab="API (2000)", main="API (2000) Scores")
```

... especially if we want to display variables divided by a category: 
```{r}
#| fig.width=5,
#| fig.align="center",
#| fig.cap="Distribution of API (2000) sample data"
svyboxplot(api00~stype, design=apiclus2.des, 
           xlab="School Type",
           ylab="API (2000)", main="API (2000) Scores")
```

If we have two continuous variables we can plot them as a scatterplot, showing
the weights of each observation:
```{r}
#| fig.width=5,
#| fig.align="center",
#| fig.cap="Distribution of API scores in 1999 and 2000"
svyplot(api00~api99, design=apiclus2.des, style="bubble",
        xlab="API (1999)", ylab="API (2000)")
```

### Estimation {-}

We've already seen that we can estimate means using the `svymean()` function.
```{r}
svymean(~api00, design=apiclus2.des)
```
We can do this for a set of separate variables, using a perhaps uninintuitive notation:
```{r}
svymean(~api99+api00, design=apiclus2.des)
```
or for a variable split by categories by wrapping `svymean()` up inside a call to `svyby()`
```{r}
svyby(~api00, by=~stype, svymean, design=apiclus2.des)
```

If we want to test hypotheses about differences between groups one way is to use the regression
framework
```{r}
summary(svyglm(api00~stype, design=apiclus2.des))
```
Here we can see that with Elementary as the reference level we 
get the same estimate of the mean API reported as the `(Intercept)`.
We further see that there is no significant difference between Middle
and Elementary Schools in mean API, but the mean API is significantly lower
at High Schools compared to Elementary Schools.

Regressions take of course take lots of variables into account, and
we can apply familiar regression techniques to investigating 
conditional associations between predictors and a continuous outcome. 

A **dotchart** is a nice way of showing the estimated means and their confidence intervals:
```{r}
#| fig.width=5,
#| fig.align="center",
#| fig.cap="Mean API (2000) scores by School Type"
ss <- svyby(~api00, by=~stype, svymean, design=apiclus2.des)
sc <- confint(ss)
ns <- nrow(sc)
dotchart(ss, xlim=range(sc), xlab="Mean API (2000) Score")
arrows(sc[,1], 1:ns, sc[,2], 1:ns, angle=90, code=3, length=0.1)
```

We can estimate totals using the `svytotal()` function - for example
we can estimate the total number of students enrolled at all schools in 
California:
```{r}
svytotal(~enroll, design=apiclus2.des, na.rm=TRUE)
```
Although note that we have had to remove some schools with unknown
`enroll` values by stating `na.rm=TRUE`, and these forced removals likely mean 
that our estimate of the total is an underestimate.

Confidence intervals can be calculated using the `confint()` function
```{r}
confint(svymean(~api99+api00, design=apiclus2.des), level=0.95)
```

```{r}
confint(svyby(~api00, by=~stype, svymean, design=apiclus2.des))
```

The standard errors can be directly extracted using `SE()`
```{r}
SE(svyby(~api00, by=~stype, svymean, design=apiclus2.des))
```
The **relative standard errors**, also known as **relative sampling errors** 
(RSEs) can be extracted by `cv()`:
```{r}
cv(svyby(~api00, by=~stype, svymean, design=apiclus2.des))
```
RSEs are a useful measure of quality of estimates of quantities that are 
strictly positive, such as the API score.

```{r echo=FALSE}
ss <- svymean(~api00, design=apiclus2.des)
cc <- cv(ss)
```

If we learned that the SE of an estimate (e.g. the mean of `api00`) was 
`r round(as.numeric(SE(ss)),2)` we wouldn't be able to assess whether it 
was a good estimate or not without knowing that the value of the estimate
was `r round(as.numeric(ss),2)`.  A good way to assess the quality of 
an estimate is the SE of an estimate $\widehat{T}$ as a proportion of 
the estimate $\widehat{T}$:
\[
   \bfa{RSE}{\widehat{T}} = \frac{\bfa{SE}{\widehat{T}}}{\widehat{T}}
\]
For overall mean `api00` the RSE is
\[
   \bfa{RSE}{\widehat{\bar{Y}}} 
   = \frac{`r round(as.numeric(SE(ss)),2)`}{`r round(as.numeric(ss),2)`} 
   = `r round(as.numeric(cc),3)`
\]
This is a highly precise estimate: the SE is only `r round(100*as.numeric(cc),1)`%
of the value of the estimate.

Good RSEs are only a few percent at most. Poor RSEs are larger than 10%, and anything bigger
than 20% isn't really publishable.

In general a standard deviation divided by a mean goes by the term
**coefficient of variation** (CV), which is why the R function is 
called `cv()`.
```{r}
cv(svymean(~api00, design=apiclus2.des))
```

## Categorical Variables

### Display {-}

The most fundamental

### Estimation {-}

Categorical variables give proportions
svymean()
svytotal()
svyciprop() for better confidence intervals

Inference

svychisq()
svytest()
svyranktest()
svyglm()
regTermTest()
svypredmeans()

Display

barplot()

Data display and inference

Analytic and jackknife methods

* SRSWOR, SRSWR
* Stratified designs
* Cluster designs
* PPSWR
* Complex designs
