# Weight adjustments

<!--- For HTML Only --- Equivalent code for .pdf must be included in preamble.tex -->
`r if (!knitr:::is_latex_output()) '
$\\DeclareMathOperator*{\\argmin}{argmin}$
$\\newcommand{\\var}{\\mathrm{Var}}$
$\\newcommand{\\bfa}[2]{{\\rm\\bf #1}[#2]}$
$\\newcommand{\\rma}[2]{{\\rm     #1}[#2]}$
$\\newcommand{\\estm}{\\widehat}$'`

## Calibration

```{r}
apipopc <- apipop[!is.na(apipop$enroll),]
apipopc$stype <- factor(as.character(apipopc$stype),levels=c("E","M","H"))
levels(apipopc$stype) <- c("Elementary","Middle","High")
apipopc$yr.rnd[is.na(apipopc$yr.rnd)] <- "Yes"
```


```{r}
mean(apipopc$enroll)
tapply(apipopc$enroll, apipopc$stype, mean)
table(apipopc$stype)
table(apipopc$stype)/nrow(apipopc)
```


```{r}
stype.benchmarks <- as.data.frame(table(apipopc$stype))
names(stype.benchmarks) <- c("stype","Freq")
stype.benchmarks
sum(stype.benchmarks$Freq) # 6157

yr.rnd.benchmarks <- as.data.frame(table(apipopc$yr.rnd))
names(yr.rnd.benchmarks) <- c("yr.rnd","Freq")
yr.rnd.benchmarks
sum(yr.rnd.benchmarks$Freq) # 6157
```

```{r}
set.seed(111)
api2sc <- apipopc |> 
  select.2sc(n=40, cfraction=0.30, mmin=2, clustervar="dnum") 
api2sc.des <- svydesign(ids=~dnum+snum, fpc=~bigN1+bigN2, data=api2sc) 
table(api2sc$stype)/nrow(api2sc)
svytable(~stype, design=api2sc.des)
svytable(~stype, design=api2sc.des)/sum(api2sc$weight)
svymean(~enroll, design=api2sc.des)
svyby(~enroll, by=~stype, api2sc.des, svymean)
```

```{r}
api2sc.des.ps <- postStratify(api2sc.des, strata=~stype, population=stype.benchmarks)
svymean(~enroll, api2sc.des.ps)
```

```{r}
api2sc.des.rake <- rake(api2sc.des, 
                        sample=list(~stype, ~yr.rnd), 
                        population=list(stype.benchmarks, yr.rnd.benchmarks))
svymean(~enroll, api2sc.des.rake)
```


With replacement: use weights only - gets SE badly wrong
```{r}
api2sc.des.wr <- svydesign(ids=~1, data=api2sc, weight=~weight) 
svytable(~stype, design=api2sc.des.wr)
svytable(~stype, design=api2sc.des.wr)/sum(api2sc$weight)
svymean(~enroll, design=api2sc.des.wr)
svyby(~enroll, ~stype, design=api2sc.des.wr, svymean)
```

```{r}
api2sc.des.wr.ps <- postStratify(api2sc.des.wr, strata=~stype, population=stype.benchmarks)
svymean(~enroll, api2sc.des.wr.ps)
```

So we need replicate weights

api2sc.jkdes <- as.svrepdesign(api2sc.des, type="JKn")
```{r}
api2sc.jkdes <- as.svrepdesign(api2sc.des, type="JK1")
```
Informative warning here: no fpc can apply at second stage or deeper

```{r}
names(api2sc.jkdes)
api2sc.jkdes$repweights$weights  # and n x n matrix of weights: each column is a replicate
```
Replicate Weights are $n/(n-1)=40/39=1.025641$ - drops one `dnum` out for each replicate.
These need to be multiplied by the original sampling weight to get the actual weights.

```{r}
api2sc.jkdes$pweights
api2sc$weight
```

JK1 for unstratified designs, JKn for stratified designs

```{r}
svymean(~enroll, design=api2sc.jkdes)
```

We might have been **supplied** these weights, someone else might have
created them.  In which case our dataset would look like this
```{r}
fwmat <- api2sc.jkdes$repweights$weights[api2sc.jkdes$repweights$index,]
colnames(fwmat) <- paste0("repweight",sprintf("%0.3d",1:ncol(fwmat)))
pweight <- api2sc.jkdes$pweights
fwmat <- diag(pweight)%*%fwmat
sum(pweight)
apply(fwmat,2,sum)
api2sc.withrep <- cbind(api2sc, pweight, fwmat)
```
We create the design using the probability weights `pweight`
and the set of replicate weights `repweight001-repweight040`.

```{r}
pweightcol <- match("pweight",names(api2sc.withrep)) # which column name is exactly equal to pweight
repweightcols <- grep("^repweight",names(api2sc.withrep)) # which columnnames start with repweight?
varcols <- (1:ncol(api2sc.withrep))
varcols <- varcols[!(varcols%in%c(pweightcol,repweightcols))]
n <- length(unique(api2sc$dnum))
api2sc.repdes <- svrepdesign(variables=api2sc.withrep[,varcols],
                             weights=api2sc.withrep[,pweightcol],
                             repweights=api2sc.withrep[,repweightcols],
                             combined.weights=TRUE,
                             scale=(n-1)/n,
                             type="JK1")
svymean(~enroll, api2sc.repdes)
```


```{r}
set.seed(222)
apipop |> 
  select.2sc(n=40, cfraction=0.30, mmin=2, clustervar="dnum") |> 
  svydesign(ids=~dnum+snum, fpc=~bigN1+bigN2, data=_) |> 
  svymean(~enroll, design=_)
```


```{r}
table(api2sc$stype)
```



Origin of missing data

Calibration for unit non-response

* Poststratification

Imputation for item non-response

## Replicate weights

Later in the course we'll see situations where non-response means that the 
weights from our sample design have to be adjusted (e.g. by poststratification).
When this happens the estimation methods associated with the sample designs above
no longer work, and we have to resort to approximations.

In such circumstances we often create multiple sets of weights in addition to the
adjusted weights.  These **replicate weights** are what are used to estimate the
standard errors, and they have the advantage that they can create standard errors
for a wider range of estimators than the usual means and totals.  

**Jacknife** weights are calculated by systematically deleting PSUs at the first
stage of selection, and then scaling the weights on the remaining rows so that they 
still add up to the same total.
