# Weight adjustments

<!--- For HTML Only --- Equivalent code for .pdf must be included in preamble.tex -->
`r if (!knitr:::is_latex_output()) '
$\\DeclareMathOperator*{\\argmin}{argmin}$
$\\newcommand{\\var}{\\mathrm{Var}}$
$\\newcommand{\\bfa}[2]{{\\rm\\bf #1}[#2]}$
$\\newcommand{\\rma}[2]{{\\rm     #1}[#2]}$
$\\newcommand{\\estm}{\\widehat}$'`

## Weight Calibration

At the end of a survey we may find ourselves in the undesirable situation that

 * **Purely by chance**, the sample is somewhat unbalanced: e.g. by chance we may have 
   selected many more males than females, or more large households than small ones, or
 * We may have had **unit nonresponse** - i.e. there may be selected units from
   which we have collected no data.  In the case of human subjects this may have
   been from a failure to contact the selected person, or the refusal of the 
   person to answer our questions.
   
In both cases we are in the situation that our sample, even when weighted, is
not a good representation of the population from which it is drawn.  

```{r echo=FALSE, message=FALSE, warning=FALSE}
nzpop <- read.csv("data/year-age-sex.csv")
nzpop <- nzpop |>
  filter(Year==2018)
nzpop <- nzpop[nzpop$Age>=15 & nzpop$Age<=45,]
idx <- nzpop$Age==45
nzpop$count[idx] <- (1/5)*nzpop$count[idx]
nzpop <- nzpop |>
  mutate(AgeGrp=case_when(Age<=20 ~ "15-24",
                          Age<=30 ~ "25-34",
                          Age<=45 ~ "35-45"))
nzpop <- nzpop |>
  mutate(count=round(count)) |>
  group_by(AgeGrp,Sex) |>
  summarise(count=sum(count)) |>
  as.data.frame()
nzpop <- nzpop |> rename(Gender=Sex)
nzpop$Gender <- factor(nzpop$Gender,levels=c("Male","Female"))
levels(nzpop$Gender) <- tolower(levels(nzpop$Gender))
total <- sum(nzpop$count)
total.Gender <- tapply(nzpop$count, nzpop$Gender, sum)
total.AgeGrp <- tapply(nzpop$count, nzpop$AgeGrp, sum)
total.AgeGrp <- data.frame(AgeGrp=names(total.AgeGrp),Nh=total.AgeGrp)
```

```{r echo=FALSE}
surf <- read.csv("data/surf.csv")[,1:8]
bigN <- total
n <- nrow(surf)
surf$Gender <- factor(surf$Gender, levels=c("male","female"))
surf <- surf |>
  mutate(AgeGrp=case_when(Age<=24 ~ "15-24",
                          Age<=34 ~ "25-34",
                          Age<=45 ~ "35-45"))
surf$bigN <- bigN
surf$weight <- bigN/n
n.Gender <- tapply(surf$Gender, surf$Gender, length)
sampweight <- bigN/n
```

For example consider the SURF dataset, which is a synthetic sample of 200 individuals
between the ages of 15 and 45.  The first 10 rows are shown below.  

```{r echo=FALSE}
surf |> slice_head(n=10) |>
  kbl() |>
  kable_styling(full_width=FALSE)
```

The dataset contains a mixture of numerical and categorical variables, all with very
self explanatory names (except note that `Income` and `Hours` are weekly quantities, 
and also that the income values are somewhat old!).  We'll treat it as a SRSWOR from
the New Zealand population between the ages of 15 and 45.  

The total population in this age range is $N=`r total`$, the sample size is $n=`r n`$, 
and so the weight of each sampled individual is 
\[
   w = \frac{N}{n} = {`r bigN`}{`r n`} = `r sprintf("%.1f",bigN/n)`
\]
Let's specify this simple random sampling design to R:
```{r}
surf.des <- svydesign(id=~1, fpc=~bigN, data=surf)
```
and then estimate the mean income:
```{r}
svymean(~Income, design=surf.des)
```
We should note that the mean income differs considerably between males and females.
```{r}
svyby(~Income, by=~Gender, design=surf.des, svymean)
```

```{r echo=FALSE}
svyboxplot(Income~Gender, design=surf.des, main="Income by Gender", ylab="Weekly income")
```

Let's also note the distribution of the population by age and sex, as measured
by the most recent census:
```{r echo=FALSE}
nzpoptab <- nzpop |> 
  arrange(Gender) |>
  pivot_wider(names_from="Gender", values_from="count") |>
  as.data.frame()
nzpoptab$Total <- apply(nzpoptab[,-1],1,sum)
nzpoptab <- rbind(nzpoptab, 
                  data.frame(AgeGrp="Total", rbind(apply(as.matrix(nzpoptab[,-1]),2,sum))))
nzpoptab.pct <- as.data.frame(lapply(nzpoptab,
  function(x) {
    if(is.numeric(x)) return(sprintf("%.1f%%",100*x/total)) else (return(x))
  }))
nzpoptab |>
  kbl() |>
  kable_styling(full_width=FALSE)
```

Or we can express these counts as percentages:
```{r echo=FALSE}
nzpoptab.pct |>
  kbl() |>
  kable_styling(full_width=FALSE)
```

The proportions of males and females in the population are roughly equal in the population, 
but in the sample they're not quite equal:
```{r}
svymean(~Gender, design=surf.des)
```
This has been caused by our having, by chance, sampled slightly more females than males.
```{r}
table(surf$Gender)
```

The consequence of this is that our estimate of the population mean income is 
and underestimate, due to females earning less than males, and their being 
slightly overrepresented in the sample.

One obvious solution to this situation is to slightly upweight the males in the sample, and downweight
the females, to take care of the imbalance.  Of course we only **know** about the 
imbalance because we have the census counts which tell us that the proportions of
males and females are roughly equal: otherwise we'd have no idea that there might be a problem.

These external counts (external to the survey that is) are called **population benchmarks**, 
and the weight adjustment that makes the weights match the benchmarks is called
**poststratification**.  It's 'post-' here because we're making the stratification 
(breaking into groups) **after** the survey has taken place.  Breaking the sample into 
groups **beforehand**, and taking separate samples within each group is called stratified
sampling.  But that is only possible if we know the group membership of sample members before
we sample them - i.e. if the group membership information is available on the frame.  

We have $n=`r n`$ individuals in a SRSWOR drawn from a population
of size $N=`r bigN`$, with weights $w=`r bigN`/`r n`=`r sprintf("%.1f",bigN/n)`$.  These
correctly add up to the population total
\[
   n \times w = n \times \frac{N}{n} = `r n`\times `r sprintf("%.1f",bigN/n)` = `r bigN`
\]
What we want is different weights for the males and females, so that they separately 
add up to the total numbers of males $N_1=`r total.Gender[1]`$ and the total
numbers of females $N_2=`r total.Gender[2]`$ in the population.  We can achieve this
by computing new post-stratified weights using the known benchmarks $N_h$ and sample
sizes $n_h$ in each of the post-strata, and computing new weights
\[
   w_h = \frac{N_h}{n_h}
\]
i.e. these are the weights we'd have used if we had actually carried out a
stratified sample in the first place.

```{r echo=FALSE}
pstab <- data.frame(Gender=c("male","female"),
                    SampWeight=sampweight,
                    nh=n.Gender,
                    Nh=total.Gender,
                    PostStratWeight=total.Gender/n.Gender)
rownames(pstab) <- NULL
pstab |>
  kbl() |>
  kable_styling(full_width=FALSE)
```

These new weights are now appropriately higher for males and lower for females.

```{r echo=FALSE}
Gender.benchmarks <- pstab[,c("Gender","Nh")]
```


In R we can modify the weights in the design object using the `postStratify()` function. 
We first need to create a data frame storing the benchmarks:
```{r}
Gender.benchmarks
```

```{r}
surf.des.ps <- postStratify(surf.des, strata=~Gender, population=Gender.benchmarks)
```

The orignal estimate of the proportions of males and females
```{r}
svymean(~Gender, design=surf.des)
```
has been modified to
```{r}
svymean(~Gender, design=surf.des.ps)
```
Notice that the Standard Errors here are zero: these proportions are no longer
estimates, they have been imposed by our benchmarks.

Then we can compare the orginal estimate of mean income
```{r}
svymean(~Income, surf.des)
```
with the post-stratified estimate
```{r}
svymean(~Income, surf.des.ps)
```
The estimate has been increased slightly as expected.  Also note that 
the Standard Error has been reduced: this is often the case with
post-stratified estimates: post-stratification makes the properties
of different samples differ less than they otherwise would, and thereby 
reduces our overall uncertainty.  

Non-response can make these small chance differences even worse.
Males and young people tend to have the lowest response rates in surveys.
So let's see what happens to our mean estimate if we take our sample of 200 individuals
and filter out 20% of the male respondents:
```{r}
set.seed(234)
surf.nonresp <- surf
surf.nonresp$Responds <- rbinom(nrow(surf),1,prob=ifelse(surf$Gender=="male",0.8,1.0))
surf.nonresp <- surf.nonresp[surf.nonresp$Responds==1,]
surf.nonresp$weight <- bigN/nrow(surf.nonresp)
table(surf.nonresp$Gender)
```

We've now got a lot fewer males than females.  Let's see what this does to the 
simple estimate of the population mean income:
```{r}
surf.nonresp.des <- svydesign(id=~1, fpc=~bigN, data=surf.nonresp)
svymean(~Income, design=surf.nonresp.des)
```

The introduction of the non-response has led to a reduction in our original estimate
from `r round(svymean(~Income, surf.des),2)` to `r round(svymean(~Income, surf.nonresp.des),2)`.

Post-stratifying however
```{r}
surf.nonresp.des.ps <- postStratify(surf.nonresp.des, strata=~Gender, 
                                    population=Gender.benchmarks)
```
leads to a dramatically improved estimate
```{r}
svymean(~Income, design=surf.nonresp.des.ps)
```

We know more than just the distribution of the population over Gender, we also know it over the
three age groups.  So we can post-stratify by these variables too.

```{r echo=FALSE}
AgeGrp.Gender.benchmarks <- nzpop
```

We can store the AgeGrp/Gender benchmarks in a data frame `AgeGrp.Gender.benchmarks`:
```{r echo=FALSE}
AgeGrp.Gender.benchmarks |> kbl() |> kable_styling(full_width=FALSE)
```
and then post-stratify
```{r}
surf.nonresp.des.ps2 <- postStratify(surf.nonresp.des, strata=~AgeGrp*Gender, 
                                     population=AgeGrp.Gender.benchmarks)
svymean(~Income, design=surf.nonresp.des.ps2)
```
And this is estimate is going to be even better than before.

However where should we stop?  Should we post-stratify by as many variables as we 
possibly can?

In general we need to ensure that we have enough individuals in the sample 
in every cross-classified cell (in this case AgeGroup by Gender) to be able to 
recalculate the weight.  Empty cells are going to cause us a problem, 
since we'll be discarding the population weight associated
with people of that type in the population who happen not to have turned up in the sample.

In population surveys run by Statistics NZ we'll often see post-stratification by age 
(in 5 year bands), sex, ethnicity (in large amalgamated groups), and large regions.  

### Poststratfication as a treatment for non-response {-}

A warning here: post-stratification is often used to modify weights in the presence
of non-response.  

### Raking {-}

We may not have full cross-classified benchmarks, for example we might know the
totals by AgeGrp and totals by Gender, but not totals by AgeGrp and Gender simultaneously.
In that case we can proceed iteratively: we post-stratify first by one variable, and
then by other.  Each post-stratification on one variable messes up the post-stratification
on the other.  But eventually the weights will converge to approximately match both sets
of benchmarks.  This iteration is called **raking** (like raking a lawn one way, then other other), 
or **Iterative Proportional Fitting**. 

```{r echo=FALSE}
AgeGrp.benchmarks <- total.AgeGrp
```

R implements raking in with the `rake()` function. We need a separate set of benchmarks
for AgeGrp
```{r echo=FALSE}
AgeGrp.benchmarks |> kbl() |> kable_styling(full_width=FALSE)
```
and for Gender
```{r echo-FALSE}
Gender.benchmarks |> kbl() |> kable_styling(full_width=FALSE)
```

With these we can rake the design:
```{r}
surf.nonresp.des.rake <- rake(surf.nonresp.des, 
                              sample=list(~AgeGrp, ~Gender), 
                              population=list(AgeGrp.benchmarks, Gender.benchmarks))
```
and then calculate our estimates
```{r}
svymean(~Income, design=surf.nonresp.des.rake)
```


### Another example {-}

In the Academic Performance Indicator dataset we have full population information
on types of School (`stype`=Elementary/Middle/High) and whether or not they
met their year round performance indicators (`yr.rnd`=Yes/No).  These counts
will be our benchmarks for post-stratification.

```{r echo=FALSE}
apipopc <- apipop[!is.na(apipop$enroll),]
apipopc$stype <- factor(as.character(apipopc$stype),levels=c("E","M","H"))
levels(apipopc$stype) <- c("Elementary","Middle","High")
apipopc$yr.rnd[is.na(apipopc$yr.rnd)] <- "Yes"
```

The numbers of schools by type are in `stype.benchmnarks`
```{r echo=FALSE}
stype.benchmarks <- as.data.frame(table(apipopc$stype))
names(stype.benchmarks) <- c("stype","Freq")
stype.benchmarks |> kbl() |> kable_styling(full_width=FALSE)
#sum(stype.benchmarks$Freq) # 6157
```

The number of schools by achievement of year round objectives are in `yr.rnd.benchmarks`
```{r echo=FALSE}
yr.rnd.benchmarks <- as.data.frame(table(apipopc$yr.rnd))
names(yr.rnd.benchmarks) <- c("yr.rnd","Freq")
yr.rnd.benchmarks |> kbl() |> kable_styling(full_width=FALSE)
#sum(yr.rnd.benchmarks$Freq) # 6157
```

and we also have these counts **cross-classified**, i.e. numbers of schools for 
each combination of `stype` and `yr.rnd`:
```{r echo=FALSE}
stype.yr.rnd.benchmarks <- as.data.frame(table(apipopc$stype, apipopc$yr.rnd))
names(stype.yr.rnd.benchmarks) <- c("stype","yr.rnd","Freq")
stype.yr.rnd.benchmarks |> kbl() |> kable_styling(full_width=FALSE)
```

Select a two stage cluster sample - and then make estimates of the mean school 
size (`enroll`)
```{r}
set.seed(111)
api2sc <- apipopc |> 
  select.2sc(n=40, cfraction=0.30, mmin=2, clustervar="dnum") 
api2sc.des <- svydesign(ids=~dnum+snum, fpc=~bigN1+bigN2, data=api2sc) 
table(api2sc$stype)/nrow(api2sc)
svytable(~stype, design=api2sc.des)
svytable(~stype, design=api2sc.des)/sum(api2sc$weight)
svymean(~enroll, design=api2sc.des)
svyby(~enroll, by=~stype, api2sc.des, svymean)
```

Poststratify by stype and reestimate
```{r}
api2sc.des.ps <- postStratify(api2sc.des, strata=~stype, population=stype.benchmarks)
svymean(~enroll, api2sc.des.ps)
```

Postratify by stype by yr.rnd and reestimate
```{r}
api2sc.des.ps2 <- postStratify(api2sc.des, strata=~stype+yr.rnd, 
                               population=stype.yr.rnd.benchmarks)
svymean(~enroll, api2sc.des.ps2)
```

Poststratify by stype + yr.rnd (i.e. we only have the marginals so we need to rake) 
and reestimate
```{r}
api2sc.des.rake <- rake(api2sc.des, 
                        sample=list(~stype, ~yr.rnd), 
                        population=list(stype.benchmarks, yr.rnd.benchmarks))
svymean(~enroll, api2sc.des.rake)
```

```{r echo=FALSE}
ff <- function(x) c(as.numeric(x), SE(x))
```


Compare
```{r echo=FALSE}
ss <- rbind(
  c(mean(apipopc$enroll), 0),
  ff(svymean(~enroll, api2sc.des)), 
  ff(svymean(~enroll, api2sc.des.ps)), 
  ff(svymean(~enroll, api2sc.des.ps2)), 
  ff(svymean(~enroll, api2sc.des.rake))) 
dimnames(ss) <- list(c("Cenus","2SC","2SC+PS(stype)","2SC+PS(stype x yr.rnd)","2SC+Rake(stype+yr.rnd)"),
                     c("Estimate","SE"))
ss |> kbl() |> kable_styling(full_width=FALSE)
```

## Replicate weights

For some weight adjustments we no longer have a reliable method of computing
the standard errors of our estimates.  This can occur when we have very 
complex designs.  It can also be necessary when we don't have access to the frame, 
and therefore don't have all the information we need to calculate standard errors.

In such circumstances we often create multiple sets of weights in addition to the
adjusted weights.  These **replicate weights** are what are used to estimate the
standard errors, and they have the advantage that they can create standard errors
for a wider range of estimators than the usual means and totals.  

**Jacknife** weights are calculated by systematically deleting PSUs at the first
stage of selection, and then scaling the weights on the remaining rows so that they 
still add up to the same total.
```{r}
api2sc.jkdes <- as.svrepdesign(api2sc.des, type="JK1")
```
There is an informative warning here: no fpc can apply at second stage of selection 
(nor any subsequent stages).  

(Also note that the method "JK1" is used for unstratified designs.  For stratified
designs the jackknife is specified by "JKn": we don't just drop out one PSU over 
the whole sample in each replicate, instead we drop out a PSU from every stratum.)

Here are the first 3 rows and 8 columns: each row refers to a district `dnum`
and each column is a replicate.
```{r}
api2sc.jkdes$repweights$weights[1:3,1:8]
```
Replicate Weights are $n/(n-1)=40/39=1.025641$ - drops one of the 40 districts `dnum` out for 
each replicate, setting all units in that district to zero.
These are scaling weights, and theyneed to be multiplied by the original sampling weight 
to get the actual weights.


```{r}
svymean(~enroll, design=api2sc.jkdes)
```
This gives us a better estimate of the SE.

If we are needing to post-stratify our weights to account for non-response, then
we need also to poststratify each set of replicate weights.

```{r}
api2sc.jkdes.ps <- postStratify(api2sc.jkdes, strata=~stype, population=stype.benchmarks)
svymean(~enroll, design=api2sc.jkdes.ps)
```
Here post-stratification gives us a strong reduction of the standard errors

Comparison
```{r}
ss <- rbind(
  c(mean(apipopc$enroll), 0),
  ff(svymean(~enroll, api2sc.des)), 
  ff(svymean(~enroll, api2sc.des.ps)), 
  ff(svymean(~enroll, api2sc.des.ps2)), 
  ff(svymean(~enroll, api2sc.des.rake)),
  ff(svymean(~enroll, api2sc.jkdes)),
  ff(svymean(~enroll, api2sc.jkdes.ps))
  ) 
dimnames(ss) <- list(c("Cenus","2SC","2SC+PS(stype)","2SC+PS(stype x yr.rnd)","2SC+Rake(stype+yr.rnd)",
                       "JK1","JK1+PS(stype)"),
                     c("Estimate","SE"))
ss |> kbl() |> kable_styling(full_width=FALSE)
```

Let's take a look at the replicate weights after post-stratification
```{r}
api2sc.jkdes.ps$repweights$weights[1:3,1:8]
```

We might have been **supplied** replicate weights, i.e. someone else might have
created them.  This is commonly the case when we have been granted access
to a data set collected by a government department.

In which case our dataset would look like this
```{r}
fwmat <- api2sc.jkdes.ps$repweights$weights[api2sc.jkdes.ps$repweights$index,]
colnames(fwmat) <- paste0("repweight",sprintf("%0.3d",1:ncol(fwmat)))
pweight <- api2sc.jkdes.ps$pweights
fwmat <- diag(pweight)%*%fwmat
sum(pweight)
apply(fwmat,2,sum)
api2sc.withrep.ps <- cbind(api2sc, pweight, fwmat)
```
We create the design using the probability weights `pweight`
and the set of replicate weights `repweight001-repweight040`.

```{r}
pweightcol <- match("pweight",names(api2sc.withrep.ps)) # which column name is exactly equal to pweight
repweightcols <- grep("^repweight",names(api2sc.withrep.ps)) # which columnnames start with repweight?
varcols <- (1:ncol(api2sc.withrep.ps))
varcols <- varcols[!(varcols%in%c(pweightcol,repweightcols))]
n <- length(unique(api2sc$dnum))
api2sc.repdes.ps <- svrepdesign(variables=api2sc.withrep.ps[,varcols],
                                weights=api2sc.withrep.ps[,pweightcol],
                                repweights=api2sc.withrep.ps[,repweightcols],
                                combined.weights=TRUE,
                                scale=(n-1)/n,
                                type="JK1")
svymean(~enroll, api2sc.repdes.ps)
```
```{r}
ss <- rbind(
  c(mean(apipopc$enroll), 0),
  ff(svymean(~enroll, api2sc.des)), 
  ff(svymean(~enroll, api2sc.des.ps)), 
  ff(svymean(~enroll, api2sc.des.ps2)), 
  ff(svymean(~enroll, api2sc.des.rake)),
  ff(svymean(~enroll, api2sc.jkdes)),
  ff(svymean(~enroll, api2sc.jkdes.ps)),
  ff(svymean(~enroll, api2sc.repdes.ps))
  ) 
dimnames(ss) <- list(c("Cenus","2SC","2SC+PS(stype)","2SC+PS(stype x yr.rnd)","2SC+Rake(stype+yr.rnd)",
                       "JK1","JK1+PS(stype)","JK1+PS(stype)"),
                     c("Estimate","SE"))
ss
```

JKn is used with stratified designs

E.g. draw a two stage cluster sample within two strata: schools which are or are
not eligible for an awards programme:
```{r}
set.seed(999)
apist2sc <- apipopc |>
  select.st2sc(n=40, stratumvar="awards", cfraction=0.20, clustervar="dnum")
apist2sc.des <- svydesign(id=~dnum+snum, strata=~awards, fpc=~bigN1+bigN2, 
                          data=apist2sc, nest=TRUE)
apist2sc.des.ps <- postStratify(apist2sc.des, strata=~stype, population=stype.benchmarks)
```
Note: we need `nest=TRUE` here because districts `dnum` do not nest within strata.
I.e. it is the **combination** of `awards` and `dnum` that actually defines the clusters here.

```{r}
svymean(~enroll, design=apist2sc.des)
svymean(~enroll, design=apist2sc.des.ps)
apist2sc.jkdes <- as.svrepdesign(apist2sc.des, type="JKn")
svymean(~enroll, design=apist2sc.jkdes)
apist2sc.jkdes.ps <- postStratify(apist2sc.jkdes, strata=~stype, population=stype.benchmarks)
svymean(~enroll, design=apist2sc.jkdes.ps)
```

There are alternative methods to Jacknife generation of replicate weights.
You may hear of Bootstrap weights, and Balanced Repeated Replication.  Some
of these have multiple variants.  We don't need to worry about these.

### Demonstration of ps SE improvement

```{r}
set.seed(222)
R <- 100
sim <- t(sapply(1:R, 
              function(i) {
  ss <- apipopc |> 
    select.2sc(n=100, cfraction=0.30, mmin=2, clustervar="dnum") 
  d1 <- svydesign(ids=~dnum+snum, fpc=~bigN1+bigN2, data=ss)  
  d2 <- postStratify(d1, strata=~stype, population=stype.benchmarks)
  e1 <- ff(svymean(~api00, design=d1))
  e2 <- ff(svymean(~api00, design=d2))
  return(c(e1,e2))
}))
```


```{r}
par(mfrow=c(1,2))
hist(sim[,1],main="Simple estimates",xlim=c(0.9,1.1)*range(sim[,c(1,3)]))
abline(v=mean(apipopc$api00),col="red",lwd=2)
hist(sim[,3],main="Postratified estimates",xlim=c(0.9,1.1)*range(sim[,c(1,3)]))
abline(v=mean(apipopc$api00),col="red",lwd=2)
```

```{r}
par(mfrow=c(1,2))
hist(sim[,2],main="SE of Simple estimates",xlim=c(0.8,1.2)*range(sim[,c(2,4)]))
abline(v=sd(sim[,1]),col="red",lwd=2)
hist(sim[,4],main="SE of Postratified estimates",xlim=c(0.8,1.2)*range(sim[,c(2,4)]))
abline(v=sd(sim[,3]),col="red",lwd=2)

```



