# Weight adjustments

<!--- For HTML Only --- Equivalent code for .pdf must be included in preamble.tex -->
`r if (!knitr:::is_latex_output()) '
$\\DeclareMathOperator*{\\argmin}{argmin}$
$\\newcommand{\\var}{\\mathrm{Var}}$
$\\newcommand{\\bfa}[2]{{\\rm\\bf #1}[#2]}$
$\\newcommand{\\rma}[2]{{\\rm     #1}[#2]}$
$\\newcommand{\\estm}{\\widehat}$'`

## Calibration

```{r}
apipopc <- apipop[!is.na(apipop$enroll),]
apipopc$stype <- factor(as.character(apipopc$stype),levels=c("E","M","H"))
levels(apipopc$stype) <- c("Elementary","Middle","High")
apipopc$yr.rnd[is.na(apipopc$yr.rnd)] <- "Yes"
```


```{r}
mean(apipopc$enroll)
tapply(apipopc$enroll, apipopc$stype, mean)
table(apipopc$stype)
table(apipopc$stype)/nrow(apipopc)
```


```{r}
stype.benchmarks <- as.data.frame(table(apipopc$stype))
names(stype.benchmarks) <- c("stype","Freq")
stype.benchmarks
sum(stype.benchmarks$Freq) # 6157

yr.rnd.benchmarks <- as.data.frame(table(apipopc$yr.rnd))
names(yr.rnd.benchmarks) <- c("yr.rnd","Freq")
yr.rnd.benchmarks
sum(yr.rnd.benchmarks$Freq) # 6157

stype.yr.rnd.benchmarks <- as.data.frame(table(apipopc$stype, apipopc$yr.rnd))
names(stype.yr.rnd.benchmarks) <- c("stype","yr.rnd","Freq")
stype.yr.rnd.benchmarks
```

Select a two stage cluster sample - make estimates
```{r}
set.seed(111)
api2sc <- apipopc |> 
  select.2sc(n=40, cfraction=0.30, mmin=2, clustervar="dnum") 
api2sc.des <- svydesign(ids=~dnum+snum, fpc=~bigN1+bigN2, data=api2sc) 
table(api2sc$stype)/nrow(api2sc)
svytable(~stype, design=api2sc.des)
svytable(~stype, design=api2sc.des)/sum(api2sc$weight)
svymean(~enroll, design=api2sc.des)
svyby(~enroll, by=~stype, api2sc.des, svymean)
```

Poststratify by stype and reestimate
```{r}
api2sc.des.ps <- postStratify(api2sc.des, strata=~stype, population=stype.benchmarks)
svymean(~enroll, api2sc.des.ps)
```

Postratify by stype x yr.rnd and reestimate
```{r}
api2sc.des.ps2 <- postStratify(api2sc.des, strata=~stype+yr.rnd, population=stype.yr.rnd.benchmarks)
svymean(~enroll, api2sc.des.ps2)
```

Poststratify by stype + yr.rnd (i.e. we only have the marginals) and reestimate
```{r}
api2sc.des.rake <- rake(api2sc.des, 
                        sample=list(~stype, ~yr.rnd), 
                        population=list(stype.benchmarks, yr.rnd.benchmarks))
svymean(~enroll, api2sc.des.rake)
```

```{r}
ff <- function(x) c(as.numeric(x), SE(x))
```


Compare
```{r}
ss <- rbind(
  c(mean(apipopc$enroll), 0),
  ff(svymean(~enroll, api2sc.des)), 
  ff(svymean(~enroll, api2sc.des.ps)), 
  ff(svymean(~enroll, api2sc.des.ps2)), 
  ff(svymean(~enroll, api2sc.des.rake))) 
dimnames(ss) <- list(c("Cenus","2SC","2SC+PS(stype)","2SC+PS(stype x yr.rnd)","2SC+Rake(stype+yr.rnd)"),
                     c("Estimate","SE"))
ss
```

With replacement: specify weights only - same as 2SC estimate, but it gets SE wrong
```{r}
api2sc.des.wr <- svydesign(ids=~1, data=api2sc, weight=~weight) 
svytable(~stype, design=api2sc.des.wr)
svytable(~stype, design=api2sc.des.wr)/sum(api2sc$weight)
svymean(~enroll, design=api2sc.des.wr)
svyby(~enroll, ~stype, design=api2sc.des.wr, svymean)
```

## Replicate weights

So we need replicate weights - generate these

In such circumstances we often create multiple sets of weights in addition to the
adjusted weights.  These **replicate weights** are what are used to estimate the
standard errors, and they have the advantage that they can create standard errors
for a wider range of estimators than the usual means and totals.  

**Jacknife** weights are calculated by systematically deleting PSUs at the first
stage of selection, and then scaling the weights on the remaining rows so that they 
still add up to the same total.
```{r}
api2sc.jkdes <- as.svrepdesign(api2sc.des, type="JK1")
```
Informative warning here: no fpc can apply at second stage or deeper.

```{r}
names(api2sc.jkdes)
api2sc.jkdes$repweights$weights[1:3,]  # an n x n matrix of weights: each column is a replicate, each row is a first stage unit
```
Replicate Weights are $n/(n-1)=40/39=1.025641$ - drops one `dnum` out for each replicate.
These need to be multiplied by the original sampling weight to get the actual weights.

Use JK1 for unstratified designs, JKn for stratified designs

```{r}
svymean(~enroll, design=api2sc.jkdes)
```
Better estimate of the SE.

We can poststratify these weights by the way: we poststratify each
replicate weight set in addition to the probability weights themselves.

```{r}
api2sc.jkdes.ps <- postStratify(api2sc.jkdes, strata=~stype, population=stype.benchmarks)
svymean(~enroll, design=api2sc.jkdes.ps)
```
We get immediate reduction of the standard errors

Comparison
```{r}
ss <- rbind(
  c(mean(apipopc$enroll), 0),
  ff(svymean(~enroll, api2sc.des)), 
  ff(svymean(~enroll, api2sc.des.ps)), 
  ff(svymean(~enroll, api2sc.des.ps2)), 
  ff(svymean(~enroll, api2sc.des.rake)),
  ff(svymean(~enroll, api2sc.jkdes)),
  ff(svymean(~enroll, api2sc.jkdes.ps))
  ) 
dimnames(ss) <- list(c("Cenus","2SC","2SC+PS(stype)","2SC+PS(stype x yr.rnd)","2SC+Rake(stype+yr.rnd)",
                       "JK1","JK1+PS(stype)"),
                     c("Estimate","SE"))
ss
```
Take a look at the replicate weights after post-stratification
```{r}
api2sc.jkdes.ps$repweights[1:3,]
```

We might have been **supplied** these weights, someone else might have
created them.  In which case our dataset would look like this
```{r}
fwmat <- api2sc.jkdes.ps$repweights$weights[api2sc.jkdes.ps$repweights$index,]
colnames(fwmat) <- paste0("repweight",sprintf("%0.3d",1:ncol(fwmat)))
pweight <- api2sc.jkdes.ps$pweights
fwmat <- diag(pweight)%*%fwmat
sum(pweight)
apply(fwmat,2,sum)
api2sc.withrep.ps <- cbind(api2sc, pweight, fwmat)
```
We create the design using the probability weights `pweight`
and the set of replicate weights `repweight001-repweight040`.

```{r}
pweightcol <- match("pweight",names(api2sc.withrep.ps)) # which column name is exactly equal to pweight
repweightcols <- grep("^repweight",names(api2sc.withrep.ps)) # which columnnames start with repweight?
varcols <- (1:ncol(api2sc.withrep.ps))
varcols <- varcols[!(varcols%in%c(pweightcol,repweightcols))]
n <- length(unique(api2sc$dnum))
api2sc.repdes.ps <- svrepdesign(variables=api2sc.withrep.ps[,varcols],
                                weights=api2sc.withrep.ps[,pweightcol],
                                repweights=api2sc.withrep.ps[,repweightcols],
                                combined.weights=TRUE,
                                scale=(n-1)/n,
                                type="JK1")
svymean(~enroll, api2sc.repdes.ps)
```
```{r}
ss <- rbind(
  c(mean(apipopc$enroll), 0),
  ff(svymean(~enroll, api2sc.des)), 
  ff(svymean(~enroll, api2sc.des.ps)), 
  ff(svymean(~enroll, api2sc.des.ps2)), 
  ff(svymean(~enroll, api2sc.des.rake)),
  ff(svymean(~enroll, api2sc.jkdes)),
  ff(svymean(~enroll, api2sc.jkdes.ps)),
  ff(svymean(~enroll, api2sc.repdes.ps))
  ) 
dimnames(ss) <- list(c("Cenus","2SC","2SC+PS(stype)","2SC+PS(stype x yr.rnd)","2SC+Rake(stype+yr.rnd)",
                       "JK1","JK1+PS(stype)","JK1+PS(stype)"),
                     c("Estimate","SE"))
ss
```

JKn is used with stratified designs

E.g. draw a two stage cluster sample within two strata: schools which are or are
not eligible for an awards programme:
```{r}
set.seed(999)
apist2sc <- apipopc |>
  select.st2sc(n=40, stratumvar="awards", cfraction=0.20, clustervar="dnum")
apist2sc.des <- svydesign(id=~dnum+snum, strata=~awards, fpc=~bigN1+bigN2, 
                          data=apist2sc, nest=TRUE)
apist2sc.des.ps <- postStratify(apist2sc.des, strata=~stype, population=stype.benchmarks)
```
Note: we need `nest=TRUE` here because districts `dnum` do not nest within strata.
I.e. it is the **combination** of `awards` and `dnum` that actually defines the clusters here.

```{r}
svymean(~enroll, design=apist2sc.des)
svymean(~enroll, design=apist2sc.des.ps)
apist2sc.jkdes <- as.svrepdesign(apist2sc.des, type="JKn")
svymean(~enroll, design=apist2sc.jkdes)
apist2sc.jkdes.ps <- postStratify(apist2sc.jkdes, strata=~stype, population=stype.benchmarks)
svymean(~enroll, design=apist2sc.jkdes.ps)
```

There are alternative methods to Jacknife generation of replicate weights.
You may hear of Bootstrap weights, and Balanced Repeated Replication.  Some
of these have multiple variants.  We don't need to worry about these.

## Demonstration of ps SE improvement

```{r}
set.seed(222)
R <- 100
sim <- t(sapply(1:R, 
              function(i) {
  ss <- apipopc |> 
    select.2sc(n=100, cfraction=0.30, mmin=2, clustervar="dnum") 
  d1 <- svydesign(ids=~dnum+snum, fpc=~bigN1+bigN2, data=ss)  
  d2 <- postStratify(d1, strata=~stype, population=stype.benchmarks)
  e1 <- ff(svymean(~api00, design=d1))
  e2 <- ff(svymean(~api00, design=d2))
  return(c(e1,e2))
}))
```


```{r}
par(mfrow=c(1,2))
hist(sim[,1],main="Simple estimates",xlim=c(0.9,1.1)*range(sim[,c(1,3)]))
abline(v=mean(apipopc$api00),col="red",lwd=2)
hist(sim[,3],main="Postratified estimates",xlim=c(0.9,1.1)*range(sim[,c(1,3)]))
abline(v=mean(apipopc$api00),col="red",lwd=2)
```

```{r}
par(mfrow=c(1,2))
hist(sim[,2],main="SE of Simple estimates",xlim=c(0.8,1.2)*range(sim[,c(2,4)]))
abline(v=sd(sim[,1]),col="red",lwd=2)
hist(sim[,4],main="SE of Postratified estimates",xlim=c(0.8,1.2)*range(sim[,c(2,4)]))
abline(v=sd(sim[,3]),col="red",lwd=2)

```



