# Imputation

Dealing with item non-response

Can do complete case analysis, but that risks both bias and
loss of precision due to reduced sample size.

Another option is imputation: fill in the data using a model.
All such models involve untestable assumptions.

```{r}
surf <- read.csv("data/surf.csv")[,1:8]
```

Remove some data: marital status and income
```{r}
set.seed(111)
msurf <- surf
msurf$Marital[sample(nrow(surf),10)] <- NA
msurf$Income[sample(nrow(surf),10)] <- NA
table(msurf$Marital,exclude=NULL)
sum(is.na(msurf$Income))
```

```{r message=FALSE, warning=FALSE}
library(mice)
library(lattice)
```

Mean imputation/cell mean imputation: method = 'mean'
Median imputation/cell median imputation: 
Hot deck cell imputation: method = 'sample' within cells
Regression imputation

Don't impute a column by setting the method="" for that column

'blocks' are sets of variables, treated similarly
```{r}
mice(msurf, m=5, 
     method=,
     blocks=)
```

Example dataset supplied with mice is `nhanes` - note all variables are numeric
```{r}
mnhanes <- mice::nhanes
mnhanes$age <- factor(mnhanes$age)
mnhanes <- data.frame(id=factor(1:nrow(mnhanes)),mnhanes)
dim(mnhanes) # 25 x 5
summary(mnhanes)
```

Surf data is a mixture of numeric and categorical variables
```{r}
summary(msurf)
```


Inspect missing pattern
```{r}
md.pattern(mnhanes)
md.pattern(msurf)
```

Overall mean imputation
```{r}
mnhanes.mfit1 <- mnhanes |> mice(method="mean", m=1, maxit=1)
mnhanes.m1 <- complete(mnhanes.mfit1)
md.pattern(mnhanes.m1) # no missing data left
```

Overall mean imputation
```{r}
mnhanes.grandmean <- mnhanes |>
  mice(method="mean", m=1) |>
  complete() 
```

Cell mean imputation
```{r}
mnhanes.cellmean <- do.call(rbind, 
                            by(mnhanes, mnhanes$age,
                               function(dsub) complete(mice(dsub, method="mean", m=1)))) 
mnhanes.cellmean <- arrange(mnhanes.cellmean, id)
```

Compare
```{r}
mnhanes.grandmean[is.na(mnhanes$bmi),]
mnhanes.cellmean[is.na(mnhanes$bmi),]
```

Repeat cell mean imputation using the regression formulation

```{r}
allvarnames <- names(mnhanes)

blocks <- as.list(allvarnames) # each variable in its own block
names(blocks) <- allvarnames
method <- rep("",length(allvarnames)) # method for each block: start with no imputation
names(method) <- allvarnames
formulas <- as.list(rep("~1", length(allvarnames)))
names(formulas) <- allvarnames
formulas <- lapply(formulas, formula)

method["bmi"] <- "norm.predict" # we're only going to impute bmi, by regression imputation
formulas[["bmi"]] <- formula(~age)

mnhanes.cellmeanregf <- mnhanes |> 
  mice(m=1, method=method, blocks=blocks, formulas=formulas) |>
  complete()

mnhanes.cellmeanregf
```

Compare
```{r}
mnhanes.grandmean[is.na(mnhanes$bmi),]
mnhanes.cellmean[is.na(mnhanes$bmi),]
mnhanes.cellmeanregf[is.na(mnhanes$bmi),]
```


Means of numeric and modes of categorical variables
```{r}
getmeanmode <- function(df) {
  as.data.frame(lapply(df, 
       function(x) {
         if(is.factor(x) || is.character(x)) {
           xt <- table(x)
           mval <- names(xt)[which.max(xt)]
         } else if(is.numeric(x)) {
           mval <- mean(x, na.rm=TRUE)
         } else {
           mval <- NA
         }
         return(mval)
       }))
}
```


```{r}
getmeanmode(mnhanes)
```

```{r}
sapply(names(mnhanes.m1), function(varname) mnhanes.m1[is.na(mnhanes[,varname]),varname])
```

Just do the numerical variables (gives a warning because Marital is categorical)
```{r}
msurf.mfit1 <- msurf |> mice(method="mean", m=1, maxit=1)
msurf.m1 <- complete(msurf.mfit1)
md.pattern(msurf.m1) # still has missing Marital data
```

Avoid the error
```{r}
msurf.mfit1 <- msurf[,sapply(msurf,is.numeric)] |> mice(method="mean", m=1, maxit=1)
msurf.m1 <- complete(msurf.mfit1)
md.pattern(msurf.m1) # this deletes the Marital column of course
```



Hot deck imputation
```{r}
mnhanes.hdfit1 <- mice(mnhanes, method="sample", m=1, seed=123)
mnhanes.hd1 <- complete(mnhanes.hdfit1)
mnhanes.hd1[is.na(mnhanes$bmi),]
table(mnhanes$bmi,exclude=NULL)
table(mnhanes.hd1$bmi)
```

Hot deck within cells [*** not yet done ***]
```{r}
mnhanes.hdfit1 <- mice(mnhanes, method="sample", m=1, blocks=, seed=123)
mnhanes.hd1 <- complete(mnhanes.hdfit1)
mnhanes.hd1[is.na(mnhanes$bmi),]
table(mnhanes$bmi,exclude=NULL)
table(mnhanes.hd1$bmi)
```


Regression imputation

Use regressions on all available observations?
```{r}
mnhanes.mfit2 <- mice(mnhanes, method="norm.predict", m=1, maxit=1)
mnhanes.m2 <- complete(mnhanes.mfit2)
mnhanes.m1[is.na(mnhanes$bmi),]
mnhanes.m2[is.na(mnhanes$bmi),]
```

Multiple imputation

Instead of imputing a **single** value for each missing observation, we impute
several possible imputed values - we end up with multiple completed data sets.
The variability among estimates from these separate completed data sets gives
us additional insight into the effect of our imputations, and the uncertainty
we have associated with having missing data.

The confidence we can have regarding the true scale of the uncertainty depends
on how strongly we believe in the model we've used to impute.  

Hot deck multiple imputation
```{r}
mnhanes.multfit1 <- mice(mnhanes, method="sample", m=5, seed=123)
mnhanes.mult1 <- complete(mnhanes.multfit1)
mnhanes.mult1[is.na(mnhanes$bmi),]
```

Regression mean imputation
```{r}
mice(mnhanes, method="norm.nob", m=1, maxit=1, seed=123)
```

Stochastic regression imputation: we sample from the fitted distribution
```{r}
mice(mnhanes, method="norm.nob", m=1, maxit=1, seed=123)
```

