# Imputation

In the previous chapter we saw methods of dealing with **unit non-response**: i.e.
we have units that are selected, but we collect no data at all from them.  

Another common type of non-response is where a unit provides some but not all of the
information requested: certain data items are therefore missing.  This is called
**item non-response**.

We could just do a complete case analysis - in other words throw all of the data away for 
the partially responding units, and treat them as unit non-responders with zero weight.  This wastes data and effort, 
and misses the opportunity to use the data we have collected.  It also leads to
a loss of precision due to reduced sample size.

An alternative option is **imputation**: this means that we **fill in (i.e. invent) 
the missing data** using some kind of model.

All such models involve untestable assumptions, and we run the similar risks we
saw when making weight adjustments for unit non-response.  There is a substantial risk of bias
in our estimates whatever we do: leaving these units fully out of our analysis
risks missing an important part of the population.  Incorrectly guessing their responses
is of course also a risk if we do so incorrectly.  We may be **introducing** bias 
in doing so.

Whatever we do our imputation must:

 * be **minimal** - we should not impute more than a few per cent of our data;
 * **never be done on our key variables** - if we're trying to find out about political
   views then the last thing we should be doing is guessing them.  Less significant 
   variables, e.g. household income, will have a lesser effect on our estimates;
 * be **documented** - we should always know how much imputation has been done, 
   on which variables, and most importantly **how** the imputation was done.

Let's take as an example the SURF income survey synthetic dataset of 200 individuals
aged 15-45 from the NZ population.  The top 10 rows are:
```{r echo=FALSE}
surf <- read.csv("data/surf.csv")[,1:8]
```

Let's remove some data: delete 10 entries at random from each of marital status and income.
```{r}
set.seed(111)
msurf <- surf
msurf$Marital[sample(nrow(surf),10)] <- NA
msurf$Income[sample(nrow(surf),10)] <- NA
```

Here are the counts of the various responses to the `Marital` variable after
removing 10 values.
```{r echo=FALSE}
table(msurf$Marital,exclude=NULL) |> 
  as.data.frame() |>
  rename(Marital=Var1) |>
  kbl() |> kable_styling(full_width=FALSE)
```

```{r message=FALSE, warning=FALSE}
library(mice)
library(lattice)
```

The R package `mice` provides for multiple imputation methods, and provides some 
demonstration datasets and some visualisation tools. However it's a bit tricky to use, 
so we'll use a simpler function `impute()`.

The `mice` package can draw a nice picture of the pattern of missingness in the 
SURF data, which is a mixture of numeric and categorical variables
```{r}
summary(msurf)
```

```{r}
md.pattern(msurf)
```

An example dataset supplied with the `mice` package is `nhanes`, and is a tiny 
extract of 25 rows from the US National Health and Nutrition Examination Survey
(NHANES). Note all variables are stored as numeric, though the `age` and `hyp` column 
are actually categorical.

We'll create a copy of the dataset, call it `mnhanes` and recode the factor levels.
```{r}
mnhanes <- mice::nhanes
mnhanes$age <- factor(mnhanes$age)
mnhanes$hyp <- factor(mnhanes$hyp)
levels(mnhanes$age) <- c("20-39","40-59","60+")
levels(mnhanes$hyp) <- c("No","Yes")
mnhanes <- data.frame(id=factor(1:nrow(mnhanes)),mnhanes)
```

Here are the first 10 rows of the data, which include a lot of missing
data for the variables `bmi` (body mass index, kg/m$^2$), `hyp` (presence of hypertension, 1=no, 2=yes) 
and `chl` (total serum cholesterol level, mg/dL). 
```{r echo=FALSE}
mnhanes[1:10,] |> kbl() |> kable_styling(full_width=FALSE)
```

Item non-response is indicated by an `NA` value.  Notice that units 1, 4 and 10
provided no data at all, apart from their ages.  

Here's the missingness pattern in the NHANES dataset.
```{r}
md.pattern(mnhanes)
```

## Mean, median or mode imputation

The simplest method of imputing a variable is just to replace any missing
value with a **typical** value from the units that did provide a response.

For numerical variables that can mean replacing missing values with the 
mean or the median.  (We prefer the median of course if there are outliers.)
For categorical variables the mode is the only **typical** value we have.

Let's replace the missing values of `bmi` in the NHANES `mnhanes` dataset with the
mean of the responding values:
```{r}
mnhanes1 <- mnhanes
mnhanes1$bmi.imp <- is.na(mnhanes1$bmi)
mnhanes1$bmi[is.na(mnhanes1$bmi)] <- mean(mnhanes1$bmi, na.rm=TRUE)
```

The code above does two things: in a new copy of the data set `mnhanes1` 
it creates a new column `bmi.imp` which is `TRUE` if the `bmi` value is missing.  
It then replaces `NA` values in the `bmi` column with the mean value
`r mean(mnhanes1$bmi[!mnhanes1$bmi.imp])`. 

Here are the top 10 rows of the dataset after imputation.
```{r echo=FALSE}
mnhanes1[1:10,] |> kbl() |> kable_styling(full_width=FALSE)
```


The function `impute()` does the same thing: we just specify the variable name 
and the method (mean, median or mode).
```{r}
mhnanes1 <- impute(mnhanes, varname="bmi", method="mean")
```

We can impute the categorical `hyp` variable using mode imputation 
and the numerical `chl` variable using median imputation
```{r}
mnhanes1 <- mnhanes1 |>
  impute(varname="hyp", method="mode") |>
  impute(varname="chl", method="median") 
```

This leads to the fully imputed dataset (showing the first 10 rows only): 
```{r echo=FALSE}
mnhanes1[1:10,] |> kbl() |> kable_styling(full_width=FALSE)
```

There's an obvious risk here: by filling in the
most common or most typical value in the dataset we are **reducing variability** among
the respondents and making the sample dominated by these typical values.

For example in Figure \@ref(fig:hypbarplot-mean) the proportion of people with hypertension
reduces from 4/17=24% to 4/25=16% because all of the missing individuals are assigned
to the typical reponse of 'no hypertension'.

```{r hypbarplot-mean, echo=FALSE}
#| fig.cap="Distribution of the Hypertension variable before and after mean imputation",
#| fig.width=5,
#| fig.align='center'
barplot(cbind(table(mnhanes$hyp),
              table(mnhanes1$hyp)),beside=TRUE,
        legend=TRUE, args.legend=list(x="topleft",cex=1.0),
        names=c("Before imputation","After imputation"))
```

Likewise the distribution of BMI after mean imputation is far from realistic
with a large artificial peak at the mean (Figure \@ref(fig:bmidistribution-mean)).

```{r bmidistribution-mean, echo=FALSE}
#| fig.cap="BMI distributions before and after mean imputation",
#| fig.width=8,
#| fig.align="center"
par(mfrow=c(1,2))
hist(mnhanes1$bmi[!mnhanes1$bmi.imp],breaks=10,xlim=c(15,40),
     main="BMI distribution\nbefore imputation", xlab="BMI (kg/m2)")
hist(mnhanes1$bmi,breaks=10,xlim=c(15,40),
     main="BMI distribution\nafter imputation", xlab="BMI (kg/m2)")
```

If we are only imputing a small proportion of values this shouldn't matter, 
but we'd prefer to have an imputation method that didn't introduce such 
artefacts into the data.

## Hot deck imputation

An attractive alternative to replacing an observation with the mean, median
or mode is to choose another respondent at random, and copy over their
data.  These respondents are referred to as 'donors', who 'donate' a copy of their
data to the respondent with missing data. 

The `impute()` function does this, by specifying the method "sample":
```{r}
set.seed(222)
mnhanes2 <- mnhanes |> 
  impute(varname="bmi", method="sample") |>
  impute(varname="hyp", method="sample") |>
  impute(varname="chl", method="sample")
```

The improved effect of doing so can be seen in Figures \@ref(fig:hypbarplot-sample)
(where the before and after proportions of people with hypertension are 4/17=24% and 5/20=20%) 
and \@ref(fig:bmidistribution-sample).

```{r hypbarplot-sample, echo=FALSE}
#| fig.cap="Distribution of the Hypertension variable before and after Hot Deck imputation",
#| fig.width=5,
#| fig.align='center'
barplot(cbind(table(mnhanes$hyp),
              table(mnhanes2$hyp)),beside=TRUE,
        legend=TRUE, args.legend=list(x="topleft",cex=1.0),
        names=c("Before imputation","After imputation"))
```

```{r bmidistribution-sample, echo=FALSE}
#| fig.cap="BMI distributions before and after Hot Deck imputation",
#| fig.width=8,
#| fig.align="center"
par(mfrow=c(1,2))
hist(mnhanes2$bmi[!mnhanes2$bmi.imp],breaks=10,xlim=c(15,40),
     main="BMI distribution\nbefore imputation", xlab="BMI (kg/m2)")
hist(mnhanes2$bmi,breaks=10,xlim=c(15,40),
     main="BMI distribution\nafter imputation", xlab="BMI (kg/m2)")
```

The Hot Deck method gets its name from the 'deck' (like a deck of cards) of respondents 
with complete information from which we draw at random.  'Hot' refers to the fact that
we are using respondents from the same survey as the donors. 'Cold' deck imputation uses
data from another external data source as a source of donors.

**Important note** - you'll get a different set of imputed values every time
you do a hot deck imputation - so two analysts imputing data on the same data set
will end up with different imputed values and therefore (slightly) different final estimates.
If imputation is **minimal** then this shouldn't be a problem.  

## Cell methods

We can do better still.  We know that characteristics such as BMI, hypertension
and cholesterol correlate strongly with age, and we know the age of everyone.

It makes sense to do mean, median, mode **and** hot deck imputation within 
**imputation cells**.  These cells are groups of respondents who have the same
characteristics: specified by a (set of) categorical variables.  We have to know 
which imputation cell the respondent with missing data belongs to, so we must
have complete data for the variables defining the imputation cells.

Once we have imputation cells defined we can do mean, median, mode or hot deck imputation
within the cells.

In the `impute()` function we define the cells with the right hand side of a model formula
which should only include categorical variables.  If the formula includes numerical variables
then it is unlikely that there will be anyone in the same cell as the respondent who needs a donor.
So with a variable like age we use age groups to define the cells, rather than year of age. 

Let's do cell mean imputation for BMI, cell hot deck for hypertension, and cell median for cholesterol,
all with cells defined by age group.
```{r}
set.seed(333)
mnhanes3 <- mnhanes |>
  impute(varname="bmi", method="mean", formula=~age) |>
  impute(varname="hyp", method="sample", formula=~age) |>
  impute(varname="chl", method="median", formula=~age) 
```

We can compare BMI values by age group in our original and three imputed datasets
```{r bmi-comparison, echo=FALSE}
#| fig.cap="BMI distribution by age group after various imputation methods",
#| fig.width=8,
#| fig.align="center"
par(mfrow=c(2,2))
boxplot(bmi~age, data=mnhanes, main="Original data set")
boxplot(bmi~age, data=mnhanes1, main="Mean imputation")
boxplot(bmi~age, data=mnhanes2, main="Hot deck imputation")
boxplot(bmi~age, data=mnhanes3, main="Cell mean imputation")
```

## Regression methods

A further extension of imputation methods leads us to **regression imputation**:
using the complete data respondents we fit a regression model for our variable
of interest, and then use it to predict the missing values. This works for numerical
as well as categorical predictors, and includes cell mean imputation as a special case.
As with the cell methods we need the respondents who need their values imputed
to have complete data on the predictor variables.

Let's switch to imputing the 10 missing weekly incomes in the SURF dataset `msurf`.

We first note that income depends clearly both on Gender and weekly hours worked:
```{r income-hours, echo=FALSE}
#| fig.cap="Weekly Income and Weekly Hours worked by Gender.  The solid line shows the relationship for Males and the dashed line for Females",
#| fig.width=7.5,
#| fig.align="center"
plot(msurf$Hours, msurf$Income, 
     pch=ifelse(msurf$Gender=="male",16,1),
     xlab="Weekly Hours Worked", ylab="Weekly Income")
legend("topleft", pch=c(16,1), legend=c("Male","Female"))
abline(lm(Income~Hours, data=msurf[msurf$Gender=="male",]))
abline(lm(Income~Hours, data=msurf[msurf$Gender!="male",]),lty=2)
```

Let's impute the missing values by mean imputation and 
and regression imputation.  Our regression model will be
\[ (\#eq:surf-imp-model)
   \text{Income}_i = \beta_0 + \beta_1 \text{(Hours$_i$)} 
                     + \beta_2 I(\text{Gender$_i$=Female}) 
                     + \varepsilon_i
                     \qquad \text{where $\varepsilon_i\overset{\text{iid}}{\sim} N(0,\sigma^2)$}
\]


```{r}
msurf1 <- msurf |>
  impute(varname="Income", method="mean") 
msurf2 <- msurf |>
  impute(varname="Income", method="lm.mean", formula=~Hours+Gender) 
msurf3 <- msurf |>
  impute(varname="Income", method="lm.sim", formula=~Hours+Gender) 
```

The results of the three imputation methods are shown in Figure \@ref(fig:income-hours-impute).

```{r income-hours-impute, echo=FALSE}
#| fig.cap="Weekly Income and Weekly Hours worked by Gender.  The solid line shows the relationship for Males and the dashed line for Females.  The green points are cell mean imputations, the blue points are regression mean imputations and the red points are regression simulation imputations.",
#| fig.width=7.5,
#| fig.align="center"
plot(msurf$Hours, msurf$Income, 
     pch=ifelse(msurf$Gender=="male",16,1),
     xlab="Weekly Hours Worked", ylab="Weekly Income")
legend("topleft", pch=c(16,1), legend=c("Male","Female"))
abline(lm(Income~Hours, data=msurf[msurf$Gender=="male",]))
abline(lm(Income~Hours, data=msurf[msurf$Gender!="male",]),lty=2)
datf <- msurf1[msurf1$Income.imp,]
points(datf$Hours, datf$Income, pch=ifelse(datf$Gender=="male",16,1),col="green")
datf <- msurf2[msurf1$Income.imp,]
points(datf$Hours, datf$Income, pch=ifelse(datf$Gender=="male",16,1),col="blue")
datf <- msurf3[msurf1$Income.imp,]
points(datf$Hours, datf$Income, pch=ifelse(datf$Gender=="male",16,1),col="red")
```

The cell mean imputations (shown in green on the plot) are clearly very poor - 
everyone gets the same value, and in general these values are much to high since 
our missing values correspond to workers with in general few weekly hours.

The first regression imputation has method `lm.mean` fits the model in equation \@ref(eq:surf-imp-model), 
and then predicts directly onto the fitted regression line:
\[
   \text{ImputedIncome}_i = \widehat{\beta}_0 + \widehat{\beta}_1 \text{(Hours$_i$)} 
                     + \widehat{\beta}_2 I(\text{Gender$_i$=Female}) 
\]
using the fitted regression parameters $(\widehat{\beta}_0, \widehat{\beta}_1, \widehat{\beta}_2)$.
These imputed values are shown in blue, and are exactly where we'd expect a person of the
given Gender and Hours worked to be earning.

This is very good, but in the spirit of hot deck imputation we can do even better: 
we know that observations should scatter around the the regression line, rather than lying 
directly on it.  So the second regression method draws from the error distribution
\[
   \widehat{\varepsilon}_i \sim N(0,\widehat{\sigma}^2)
\]
using the fitted standard deviation $(\widehat{\sigma})$.  So the imputed value is
the expected value plus this additional error, which bumps the prediction off the regression
line:
\[
   \text{ImputedIncome}_i = \widehat{\beta}_0 + \widehat{\beta}_1 \text{(Hours$_i$)} 
                     + \widehat{\beta}_2 I(\text{Gender$_i$=Female}) 
                     + \widehat{\varepsilon}_i
\]
These imputed values are shown in red in Figure \@ref(fig:income-hours-impute), and they
look just like real data.  To get these imputed values we change the method from
"lm.mean" to "lm.sim" to change from the mean (expected) income, to a simulated value.  


## Multiple imputation

Instead of imputing a **single** value for each missing observation, we impute
several possible imputed values - we end up with multiple completed data sets,
in something of the same spirit as the replicate weights we get with the Jackknife..
The variability among estimates from these separate completed data sets gives
us additional insight into the effect of our imputations, and the uncertainty
we have associated with having missing data.  

The confidence we can have regarding the true scale of the uncertainty depends
on how strongly we believe in the model we've used to impute.  

The `mice` package implements multiple imputation, that's its major function.  
It's more detail then we need for this course, but you should be aware that
you'll see multiple imputation methods from time to time.  

