--- 
title: "STAT395: Sampling"
author: "Richard Arnold et al."
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Sampling notes."
output: 
  bookdown::html_document2: default
  bookdown::pdf_book: 
    documentclass: krantz
    latex_engine: xelatex
    keep_tex: true
    includes:
      in_header: preamble.tex
  bookdown::pdf_document2: 
    documentclass: krantz
    latex_engine: xelatex
    keep_tex: true
    includes:
      in_header: preamble.tex
---

<!--- For HTML Only --- Equivalent code for .pdf must be included in preamble.tex -->
`r if (!knitr:::is_latex_output()) '
$\\DeclareMathOperator*{\\argmin}{argmin}$
$\\newcommand{\\var}{\\mathrm{Var}}$
$\\newcommand{\\bfa}[2]{{\\rm\\bf #1}[#2]}$
$\\newcommand{\\rma}[2]{{\\rm     #1}[#2]}$'`

```{r echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.align="center")
```

```{r warning=FALSE, error=FALSE}
library(haven)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(survey)
library(srvyr)
library(sampling)
#library(gtsummary) # error installing/loading this
library(mice)
#library(simputation) # error installing/loading this
library(scales)
```

# Introductory Remarks

These are notes for the Sampling section of STAT395/455.  The course
overlaps to some extent with STAT392/439, but there is a greater 
emphasis on R coding, computation, summary and display. 

## Lecture Plan

Course Learning Objectives

-	Describe the steps in the survey process, including the sources of survey error;
-	Draw samples from a sample frame under a variety of sample designs, and compute survey weights;
-	Carry out, interpret and present statistical inference procedures on survey data, including adjustments for non-response.

Lecture plan (3 lectures + 1 tutorial per week)

Bring a laptop to the Friday (tutorial) lecture.

### Week 1 – The Survey Process {-}

* The survey process, sources of survey error
* Populations and Frames
* Sampling in finite and infinite populations;
  Basics of SRSWOR inference; sample weights;
* **Tutorial:** Sampling basics; Population and Frames

### Week 2 – Sampling {-}

Use of `sample()` and other simulation methods in R for a variety of designs 
[`sampler` package (does power calculations, allocation, SRS and STSRS, but not cluster)] 
[`sampling` package – can do cluster samples]

*	Simple random sampling, with and without replacement
* Stratified sampling
* Cluster sampling
* Sampling in complex designs
* Sample size calculation and allocation
* Survey weights – create estimates from weights without standard errors
* **Tutorial:** using R to draw samples and compute survey weights

### Week 3 - Inference with the R survey package {-}

* Inference in simple and complex designs, including regression
* Variance estimation 
* Definition and interpretation of the design effect
* **Tutorial:** using R to specify designs and draw inferences

### Week 4 – Non-response and complex designs {-}

*	Post-stratification for non-response adjustment
* Imputation (`simputation` package)
*	Jackknife and bootstrap inference
* **Tutorial:** raking adjustments to weights; imputation; jackknife inference

### Assignment 4 (10%) [due end of Week 12: covers material from Weeks 9-11] {-}

*	Survey process, populations and frames
*	Describe a real survey, comment on design
*	Propose a sample design, frame, comment on scope and coverage
*	Draw and analyse a sample under various designs
*	Compute sample sizes and allocate samples in stratified sampling
*	Compute and interpret the design effect
* Analyse data and present results from a survey

### Test 2 (30%) [assessment period] [1.5 hours, closed book] {-}

*	Survey process, populations and frames
* Describe sample designs
* Compute and interpret sample weights
* Use sample weights for point estimates
* Interpret output from survey inference in R
* Compare results, interpret design effect
* Describe methods for non-response adjustment (imputation and post-stratification); comment on bias reduction and assumptions being made
* Describe the jackknife method of variance estimation

## Recommended reading

There are lots of books on sample surveys.

 * Thomas Lumley's book 'Complex Surveys' (@Lumley2010) is a good introduction to sample survey analysis with R,
   and accompanies the `survey` package (@survey2004).
 * Groves et al. 'Survey Methodology' (groves.etal) is a good general summary of the important issues in survey design.
 * Sarndal, Swensson and Wretman is a classic in the field of the theory of sampling (@ssw)
 * Likewise the books by @cochran and @kish
 * Little and Rubin have a great book on missing data (@littlerubin)
 * More recent are the books by @lohr and @scheaffer.mendenhall.ott
 * And in 1994 StatsNZ published 'A Guide to Good Survey Design' (@statsnz)

And also
 
 * CRAN site on R resources for Official Statistics https://cran.r-project.org/web/views/OfficialStatistics.html
 * Online book on using R for survey analysis: https://bookdown.org/jimr1603/Intermediate_R_-_R_for_Survey_Analysis/
 * This gallery of R graphs https://r-graph-gallery.com/
 
Some simple lectures on Sample Surveys

 * https://r-project.ro/conference2017/presentations/Camelia_Goga_sondages.pdf
 * Notes for epidemiologists: https://www.epirhandbook.com/en/new_pages/survey_analysis.html
 * Notes on imputation: https://www.r-bloggers.com/2023/01/imputation-in-r-top-3-ways-for-imputing-missing-data/
 
Data

 * UC Irvine Machine Learning data repository: http://archive.ics.uci.edu/
 * Some free microdata sets: https://asdfree.com/
 * Kaggle data repository: https://www.kaggle.com/
 
## R Packages

* `survey` package (@survey2004) with this handy summary of command uses https://stats.oarc.ucla.edu/r/seminars/survey-data-analysis-with-r/
* `srvyr` a dplyr wrapper for the `survey` package
* `gtsummary` creates outputs ready for publication (difficult to build/load)
* `sampling` package for drawing samples according to various designs
* `mice` package for imputation
* `simputation` package for imputation (difficult to build/load)
* `haven` package to import `SAS`, `Stata` and `SPSS` data files
 
```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'survey'
), 'packages.bib')
```

<!--- For HTML Only --- Equivalent code for .pdf must be included in preamble.tex -->
`r if (!knitr:::is_latex_output()) '
$\\DeclareMathOperator*{\\argmin}{argmin}$
$\\newcommand{\\var}{\\mathrm{Var}}$
$\\newcommand{\\bfa}[2]{{\\rm\\bf #1}[#2]}$
$\\newcommand{\\rma}[2]{{\\rm     #1}[#2]}$'`
