--- 
title: "STAT392: Sample Surveys"
author: "Richard Arnold et al."
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: book
bibliography: [book.bib, packages.bib]
biblio-style: apalike
link-citations: yes
description: "Sample Surveys notes."
output: 
  bookdown::html_document2: default
  bookdown::pdf_book: 
    documentclass: krantz
    latex_engine: xelatex
    keep_tex: true
    includes:
      in_header: preamble.tex
  bookdown::pdf_document2: 
    documentclass: krantz
    latex_engine: xelatex
    keep_tex: true
    includes:
      in_header: preamble.tex
---


```{r echo=FALSE}
knitr::opts_chunk$set(echo = TRUE, out.align="center")
```

# Introductory Remarks

These STAT 392 Lecture Notes are derived from notes created by Richard Arnold, 
Alistair Gray, and other present and former members of staff at Victoria 
University of Wellington and at Statistics New Zealand.

They should not be distributed without persmission.  

## Recommended reading

There are lots of books on sample surveys.

 * Thomas Lumley's book 'Complex Surveys' (@Lumley2010) is a good introduction to sample survey analysis with R,
   and accompanies the `survey` package (@survey2004).
 * Groves et al. 'Survey Methodology' (groves.etal) is a good general summary of the important issues in survey design.
 * Sarndal, Swensson and Wretman is a classic in the field of the theory of sampling (@ssw)
 * Likewise the books by @cochran and @kish
 * Little and Rubin have a great book on missing data (@littlerubin)
 * More recent are the books by @lohr and @scheaffer.mendenhall.ott
 * And in 1994 StatsNZ published 'A Guide to Good Survey Design' (@statsnz)
 
```{r include=FALSE}
# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown', 'survey'
), 'packages.bib')
```

<!--- For HTML Only --- Equivalent code for .pdf must be included in preamble.tex -->
`r if (!knitr:::is_latex_output()) '
$\\DeclareMathOperator*{\\argmin}{argmin}$
$\\newcommand{\\var}{\\mathrm{Var}}$
$\\newcommand{\\bfa}[2]{{\\rm\\bf #1}[#2]}$
$\\newcommand{\\rma}[2]{{\\rm     #1}[#2]}$'`
