### Assignment 4 (10%) [due end of Week 12: covers material from Weeks 9-11] {-}

*	Survey process, populations and frames
*	Describe a real survey, comment on design
*	Propose a sample design, frame, comment on scope and coverage
*	Draw and analyse a sample under various designs
*	Compute sample sizes and allocate samples in stratified sampling
*	Compute and interpret the design effect
* Analyse data and present results from a survey

### Test 2 (25%) [assessment period] [1.5 hours, closed book] {-}

*	Survey process, populations and frames
* Describe sample designs
* Compute and interpret sample weights
* Use sample weights for point estimates
* Interpret output from survey inference in R
* Compare results, interpret design effect
* Describe methods for non-response adjustment (imputation and post-stratification); comment on bias reduction and assumptions being made
* Describe the jackknife method of variance estimation


  bookdown::pdf_book: 
    documentclass: krantz
    latex_engine: xelatex
    keep_tex: true
    includes:
      in_header: preamble.tex
  bookdown::pdf_document2: 
    documentclass: krantz
    latex_engine: xelatex
    keep_tex: true
    includes:
      in_header: preamble.tex


****

Add this line to 90-BackMatter to reinstate the appendix switch:

# (APPENDIX) &nbsp; {-}

****

```{r}
pop <- read.csv("data/population.csv")
pop <- pop |> filter(Year==2023) |> 
  rename(Country=Entity, Population=all.years) |>
  arrange(Country) 
pop <- pop[!grepl("^OWID_",pop$Code),]
pop <- pop[!grepl("^UN_",pop$Code),]
pop <- pop[pop$Code!="",]
pop$Prob <- pop$Population/sum(pop$Population)
row.names(pop) <- NULL

gdp <- read.csv("data/government-expenditure-vs-gdp.csv")
gdp <- gdp |> filter(Year==2023) |> 
  rename(Country=Entity, Region="World.region.according.to.OWID") |>
  rename(GovtSpend="Government.expenditure") |>
  arrange(Country) 
gdp <- gdp[!grepl("^OWID_",pop$Code),]
gdp <- gdp[!grepl("^UN_",pop$Code),]
gdp <- gdp[gdp$Code!="",]
row.names(gdp) <- NULL

gdp <- merge(gdp,pop)
gdp$Prob <- gdp$Population/sum(gdp$Population)
gdp$GDPPerCapita <- gdp$GDP/gdp$Population

mil <- read.csv("data/military-spending-sipri.csv")
#mil <- read.csv("data/military-expenditure-as-a-share-of-government-spending.csv")
mil <- mil |> filter(Year==2023) |> 
  rename(Country=Entity, Region="World.region.according.to.OWID") |>
  rename(MilSpend=Military.expenditure) |>
  select(-Region) |>
  arrange(Country) 
mil <- mil[!grepl("^OWID_",pop$Code),]
mil <- mil[!grepl("^UN_",pop$Code),]
mil <- mil[pop$Code!="",]
row.names(mil) <- NULL

mil <- merge(gdp,mil)
#write.csv(mil, ,row.names=FALSE)
```

### Demonstration of ps SE improvement

```{r}
set.seed(222)
R <- 100
sim <- t(sapply(1:R, 
              function(i) {
  ss <- apipopc |> 
    select.2sc(n=100, cfraction=0.30, mmin=2, clustervar="dnum") 
  d1 <- svydesign(ids=~dnum+snum, fpc=~bigN1+bigN2, data=ss)  
  d2 <- postStratify(d1, strata=~stype, population=stype.benchmarks)
  e1 <- ff(svymean(~api00, design=d1))
  e2 <- ff(svymean(~api00, design=d2))
  return(c(e1,e2))
}))
```


```{r}
par(mfrow=c(1,2))
hist(sim[,1],main="Simple estimates",xlim=c(0.9,1.1)*range(sim[,c(1,3)]))
abline(v=mean(apipopc$api00),col="red",lwd=2)
hist(sim[,3],main="Postratified estimates",xlim=c(0.9,1.1)*range(sim[,c(1,3)]))
abline(v=mean(apipopc$api00),col="red",lwd=2)
```

```{r}
par(mfrow=c(1,2))
hist(sim[,2],main="SE of Simple estimates",xlim=c(0.8,1.2)*range(sim[,c(2,4)]))
abline(v=sd(sim[,1]),col="red",lwd=2)
hist(sim[,4],main="SE of Postratified estimates",xlim=c(0.8,1.2)*range(sim[,c(2,4)]))
abline(v=sd(sim[,3]),col="red",lwd=2)

```

***********


```{r}
dd <- impute(msurf, varname="Income", method="lm.mean", formula=~1)
dd[dd$Income.imp,]
mean(msurf$Income, na.rm=TRUE)
mean(dd$Income)
sd(msurf$Income, na.rm=TRUE)
sd(dd$Income)

dd <- impute(msurf, varname="Income", method="lm.sim", formula=~1)
dd[dd$Income.imp,]
mean(msurf$Income, na.rm=TRUE)
mean(dd$Income)
sd(msurf$Income, na.rm=TRUE)
sd(dd$Income)

dd <- impute(msurf, varname="Income", method="lm.mean", formula=~Gender)
dd[dd$Income.imp,]
tapply(msurf$Income, msurf$Gender, mean, na.rm=TRUE)
tapply(dd$Income, dd$Gender, mean)
tapply(msurf$Income, msurf$Gender, sd, na.rm=TRUE)
tapply(dd$Income, dd$Gender, sd)

dd <- impute(msurf, varname="Income", method="lm.sim", formula=~Gender)
dd[dd$Income.imp,]
tapply(msurf$Income, msurf$Gender, mean, na.rm=TRUE)
tapply(dd$Income, dd$Gender, mean)
tapply(msurf$Income, msurf$Gender, sd, na.rm=TRUE)
tapply(dd$Income, dd$Gender, sd)
```





Overall mean imputation
```{r}
mnhanes.mfit1 <- mnhanes |> mice(method="mean", m=1, maxit=1)
mnhanes.m1 <- complete(mnhanes.mfit1)
md.pattern(mnhanes.m1) # no missing data left
```

Overall mean imputation
```{r}
mnhanes.grandmean <- mnhanes |>
  mice(method="mean", m=1) |>
  complete() 
```

Cell mean imputation
```{r}
mnhanes.cellmean <- do.call(rbind, 
                            by(mnhanes, mnhanes$age,
                               function(dsub) complete(mice(dsub, method="mean", m=1)))) 
mnhanes.cellmean <- arrange(mnhanes.cellmean, id)
```

Compare
```{r}
mnhanes.grandmean[is.na(mnhanes$bmi),]
mnhanes.cellmean[is.na(mnhanes$bmi),]
```

Repeat cell mean imputation using the regression formulation

```{r}
allvarnames <- names(mnhanes)

blocks <- as.list(allvarnames) # each variable in its own block
names(blocks) <- allvarnames
method <- rep("",length(allvarnames)) # method for each block: start with no imputation
names(method) <- allvarnames
formulas <- as.list(rep("~1", length(allvarnames)))
names(formulas) <- allvarnames
formulas <- lapply(formulas, formula)

method["bmi"] <- "norm.predict" # we're only going to impute bmi, by regression imputation
formulas[["bmi"]] <- formula(~age)

mnhanes.cellmeanregf <- mnhanes |> 
  mice(m=1, method=method, blocks=blocks, formulas=formulas) |>
  complete()

mnhanes.cellmeanregf
```

Compare
```{r}
mnhanes.grandmean[is.na(mnhanes$bmi),]
mnhanes.cellmean[is.na(mnhanes$bmi),]
mnhanes.cellmeanregf[is.na(mnhanes$bmi),]
```


Means of numeric and modes of categorical variables
```{r}
getmeanmode <- function(df) {
  as.data.frame(lapply(df, 
       function(x) {
         if(is.factor(x) || is.character(x)) {
           xt <- table(x)
           mval <- names(xt)[which.max(xt)]
         } else if(is.numeric(x)) {
           mval <- mean(x, na.rm=TRUE)
         } else {
           mval <- NA
         }
         return(mval)
       }))
}
```


```{r}
getmeanmode(mnhanes)
```

```{r}
sapply(names(mnhanes.m1), function(varname) mnhanes.m1[is.na(mnhanes[,varname]),varname])
```

Just do the numerical variables (gives a warning because Marital is categorical)
```{r}
msurf.mfit1 <- msurf |> mice(method="mean", m=1, maxit=1)
msurf.m1 <- complete(msurf.mfit1)
md.pattern(msurf.m1) # still has missing Marital data
```

Avoid the error
```{r}
msurf.mfit1 <- msurf[,sapply(msurf,is.numeric)] |> mice(method="mean", m=1, maxit=1)
msurf.m1 <- complete(msurf.mfit1)
md.pattern(msurf.m1) # this deletes the Marital column of course
```



Hot deck imputation
```{r}
mnhanes.hdfit1 <- mice(mnhanes, method="sample", m=1, seed=123)
mnhanes.hd1 <- complete(mnhanes.hdfit1)
mnhanes.hd1[is.na(mnhanes$bmi),]
table(mnhanes$bmi,exclude=NULL)
table(mnhanes.hd1$bmi)
```

Hot deck within cells [*** not yet done ***]
```{r}
mnhanes.hdfit1 <- mice(mnhanes, method="sample", m=1, blocks=, seed=123)
mnhanes.hd1 <- complete(mnhanes.hdfit1)
mnhanes.hd1[is.na(mnhanes$bmi),]
table(mnhanes$bmi,exclude=NULL)
table(mnhanes.hd1$bmi)
```


Regression imputation

Use regressions on all available observations?
```{r}
mnhanes.mfit2 <- mice(mnhanes, method="norm.predict", m=1, maxit=1)
mnhanes.m2 <- complete(mnhanes.mfit2)
mnhanes.m1[is.na(mnhanes$bmi),]
mnhanes.m2[is.na(mnhanes$bmi),]
```





Hot deck multiple imputation
```{r}
mnhanes.multfit1 <- mice(mnhanes, method="sample", m=5, seed=123)
mnhanes.mult1 <- complete(mnhanes.multfit1)
mnhanes.mult1[is.na(mnhanes$bmi),]
```

Regression mean imputation
```{r}
mice(mnhanes, method="norm.nob", m=1, maxit=1, seed=123)
```

Stochastic regression imputation: we sample from the fitted distribution
```{r}
mice(mnhanes, method="norm.nob", m=1, maxit=1, seed=123)
```


# Package mice

Don't impute a column by setting the method="" for that column

'blocks' are sets of variables, treated similarly
```{r}
mice(msurf, m=5, 
     method=,
     blocks=)
```

Mean imputation/cell mean imputation: method = 'mean'
Median imputation/cell median imputation: 
Hot deck cell imputation: method = 'sample' within cells
Regression imputation



Get R chunks
```{r}
fnames <- list.files("../")
fnames <- fnames[grep(".Rmd$",fnames)]
for(fname in fnames) {
  stem <- gsub(".Rmd$","",fname)
  filename <- paste0("../",fname)
  outfile <- paste0(stem,"-code.R")
  print(outfile)
  x <- readLines(filename)
  x <- x[!substring(x,1,2)=="#|"]
  istart <- substring(x,1,5)=="```{r"
  iend <- substring(x,1,3)=="```" & !istart
  
  n <- length(x)
  sum(istart)==sum(iend)
  istart <- (1:n)[istart]+1
  iend <- (1:n)[iend]-1
  m <- length(istart)
  
  if(m>0) {
    sink(outfile)
    for(j in 1:m) {
      cat("\n")
      for(i in (istart[j]):(iend[j])) {
        cat(x[i]); 
        cat("\n")
      }
      cat("\n")
      cat("\n")
    }
    sink()
  }
}
```


